---
title: "RIPPLE benchmarks: FDR calibration, power, and runtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE benchmarks: FDR calibration, power, and runtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
library(ripple)
library(data.table)
library(ggplot2)
library(patchwork)
```

This vignette presents the synthetic benchmarks that validate RIPPLE's
statistical properties and runtime profile:

1. **FDR calibration under the null.** Empirical type-I error control
   with the strict gate (sign consistency = 1.0). Shows the method is
   well calibrated: false positives stay at or below the nominal rate.
2. **Relaxed-gate calibration cost.** Same setup with sign consistency
   ≥ 0.75 (the recommended setting for N ≥ 5). Shows how much
   calibration is given up when the gate is relaxed.
3. **Power curves.** Sensitivity as a function of effect size, sample
   count, and gate strictness. Shows the smallest gradient the method
   can detect at each sample size.
4. **Runtime and memory.** Wall time and peak RAM across three dataset
   sizes. Shows how the method scales.

All four panels appear in **paper Figure 2**, rendered by
`paper/render_figures.R`. This vignette uses the same cached RDS
results from `inst/extdata/`.

This vignette covers synthetic benchmarks only, so contamination
flagging does not arise here. RIPPLE never decides what counts as
contamination on its own: it proposes candidates (the cross-cell-type
`classify_gene_specificity` "broad" flag, ambient-family helpers, and
any `query_signature_genes` you supply), and you curate the final list
using domain knowledge. See the getting-started and enrichment
vignettes for how that works on real data.

Confounder control, permutation testing, fGSEA enrichment, and
ligand-receptor integration are all optional downstream steps, not
required to run the core gradient test. The other vignettes cover each
of them.

## Synthetic data design

Each dataset is a small `SpatialExperiment` with three cell types:

- **Tumor** (30 cells): dense central cluster, radius 60 µm.
- **T cells** (120 cells): uniformly distributed; the target.
- **Fibroblasts** (50 cells): annular ring at 100 to 200 µm from tumor.

For **null** datasets, all genes follow a flat Poisson rate with no
distance dependence. For **gradient** datasets, a subset of T-cell
genes have expression rates modulated by `exp(beta * distance_to_tumor)`,
where `beta` controls magnitude and direction.

## 1. FDR calibration under the null (strict gate)

500 background genes per dataset, 50 iterations per sample size
(N = 3, 5, 10), no planted gradient. The strict sign-consistency gate
(sign = 1.0) requires unanimous direction across replicates.

```{r fdr-strict, fig.cap = "Empirical FDR under a true null. Dashed line: nominal 0.05 level."}
null_path <- system.file("extdata/bench_null_hires_results.rds",
                         package = "ripple")
null_res <- readRDS(null_path)$per_run

p1 <- ggplot(null_res, aes(x = factor(n_samples), y = empirical_fdr)) +
  geom_boxplot(fill = "#E8E8E8", outlier.size = 1) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "#D55E00") +
  labs(
    x = "Number of biological replicates (N)",
    y = "Empirical FDR (Fisher FDR < 0.05)",
    title = "FDR calibration under the null (strict gate, sign = 1.0)"
  ) +
  theme_ripple(base_size = 12)
print(p1)

# Pooled-rate summary
summary_dt <- null_res[, .(
  iterations  = .N,
  mean_fdr    = sprintf("%.4f", mean(empirical_fdr)),
  max_fdr     = sprintf("%.4f", max(empirical_fdr)),
  pooled_rate = sprintf("%.4e", sum(n_sig_fdr) / sum(n_genes_tested))
), by = n_samples]
knitr::kable(
  summary_dt,
  col.names = c("N", "Iterations", "Mean FDR", "Max FDR",
                "Pooled FDR (genes)"),
  caption = "Null FDR calibration with strict gate."
)
```

The empirical FDR is near zero across all sample sizes (mean FDR
8e-5 at N = 3, 4e-5 at N = 5, 0 at N = 10). The strict gate adds a
multiplicative safety factor on top of Fisher's already-calibrated
p-value combination. Under the null, the probability that all N
per-sample coefficients agree on sign is `2 * 0.5^N`, so the gate
scales the false-positive rate by this factor before FDR correction
is even applied. Paper Figure 2A overlays this theoretical floor on
the empirical points.

## 2. Relaxed-gate calibration cost

The strict gate is conservative. Relaxing to sign consistency ≥ 0.75
(allowing 1 dissenting replicate at N = 5, 2 at N = 10) recovers power
at the cost of a higher false-positive floor. The relaxed-gate
benchmark validates that the cost is bounded.

```{r fdr-relaxed}
null_relaxed <- readRDS(system.file(
  "extdata/bench_null_hires_relaxed_results.rds",
  package = "ripple"
))$per_run

summary_relaxed <- null_relaxed[, .(
  iterations  = .N,
  mean_fdr    = sprintf("%.4f", mean(empirical_fdr)),
  max_fdr     = sprintf("%.4f", max(empirical_fdr)),
  pooled_rate = sprintf("%.4e", sum(n_sig_fdr) / sum(n_genes_tested))
), by = n_samples]
knitr::kable(
  summary_relaxed,
  col.names = c("N", "Iterations", "Mean FDR", "Max FDR",
                "Pooled FDR (genes)"),
  caption = "Null FDR calibration with relaxed gate (sign >= 0.75)."
)
```

The pooled raw-p rate rises under the relaxed gate (for example at
N = 10 it goes from 8e-5 strict to 5e-3 relaxed; see paper Figure 2B),
but the **FDR rate stays close to strict** (mean FDR is still 0 at
N = 10). BH correction absorbs most of the gate change. Relaxing the
gate is cheap on FDR but loosens raw p-value interpretation, so
downstream analysis should always use `fisher_fdr`, not raw p-values.

## 3. Power curves (strict vs relaxed gate)

5 gradient genes (negative coefficient, induced near tumor) and 495
background genes per dataset. Four effect sizes (`|beta|` = 0.002, 0.005,
0.01, 0.02) × three sample sizes (N = 3, 5, 10) × 30 iterations each.

```{r power-curves, fig.width = 8, fig.height = 5}
power_strict  <- readRDS(system.file(
  "extdata/bench_power_results.rds", package = "ripple"
))$summary
power_relaxed <- readRDS(system.file(
  "extdata/bench_power_relaxed_results.rds", package = "ripple"
))$summary

power_combined <- rbindlist(list(
  cbind(power_strict,  gate = "Strict (sign = 1.0)"),
  cbind(power_relaxed, gate = "Relaxed (sign >= 0.75)")
), use.names = TRUE)
power_combined[, gate := factor(gate,
  levels = c("Strict (sign = 1.0)", "Relaxed (sign >= 0.75)"))]

ggplot(
  power_combined,
  aes(
    x = abs(beta), y = mean_power,
    colour = factor(n_samples), linetype = gate,
    group = interaction(n_samples, gate)
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10(labels = scales::label_number()) +
  scale_colour_brewer(palette = "Set1", name = "N") +
  scale_linetype_manual(values = c(
    "Strict (sign = 1.0)"    = "solid",
    "Relaxed (sign >= 0.75)" = "dashed"
  ), name = "Gate") +
  labs(
    x = expression("|gradient score " * beta * "| (per " * mu * "m)"),
    y = "Power (sensitivity)",
    title = "Power to detect planted gradient genes"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "right", legend.box = "vertical")
```

Headline numbers (matching paper Figure 2C):

- At `|beta| >= 0.005` (a typical biological gradient), power is ~100 %
  for N >= 5 under both gates.
- At `|beta| = 0.002` (the smallest tested), the relaxed gate gains
  +5 percentage points at N = 5 and +14 pp at N = 10 vs strict.
- At N = 3 the gates coincide exactly (sign >= 0.75 of 3 ≡ sign = 1.0
  in this implementation).

```{r power-table}
knitr::kable(
  power_strict[, .(
    beta = sprintf("%.3f", beta), N = n_samples,
    power = sprintf("%.2f", mean_power),
    sd    = sprintf("%.2f", sd_power),
    fdr   = sprintf("%.3f", mean_fdr)
  )],
  col.names = c("|beta|", "N", "Power", "SD", "Observed FDR"),
  caption = "Power and observed FDR by effect size and sample count (strict gate)."
)
```

## 4. Runtime and memory

End-to-end `run_ripple()` wall time and peak resident memory across
three synthetic dataset sizes.

```{r runtime, fig.width = 7, fig.height = 4}
rt <- readRDS(system.file("extdata/bench_runtime_results.rds",
                          package = "ripple"))$per_run

rt_summary <- rt[, .(
  median_t  = median(t_ripple),
  median_mb = median(peak_mb)
), by = .(size, n_cells, n_genes, n_samples)]
rt_summary[, size := factor(size, levels = c("small", "medium", "large"))]
setorder(rt_summary, size)

knitr::kable(
  rt_summary[, .(
    size, cells = n_cells, genes = n_genes, N = n_samples,
    `time (s)`     = sprintf("%.1f", median_t),
    `peak RAM (MB)` = sprintf("%.0f", median_mb)
  )],
  caption = "Median runtime and peak RAM per dataset size (3 reps each)."
)

rt_long <- melt(
  rt_summary, id.vars = "size",
  measure.vars = c("median_t", "median_mb"),
  variable.name = "metric", value.name = "value"
)
rt_long[, metric := fifelse(metric == "median_t",
                            "Time (s)", "Peak RAM (MB)")]

ggplot(rt_long, aes(x = size, y = value, fill = metric)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = round(value)),
            position = position_dodge(width = 0.75),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c(
    "Time (s)"      = "#6A3D9A",
    "Peak RAM (MB)" = "#FF7F00"
  ), name = NULL) +
  scale_y_log10() +
  labs(
    x = "Dataset size", y = "value (log scale)",
    title = "RIPPLE runtime and peak memory"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

## Reproducing paper Figure 2

The exact 4-panel figure used in the paper (with Wilson CIs, theoretical
floors, and zero-observation triangles) is rendered by:

```{r render-fig2, eval = FALSE}
source("paper/render_figures.R")
# writes paper/figures/fig_2_stat_validation.{pdf,png}
```

## Re-running the benchmarks from scratch

The benchmark scripts live under `data-raw/benchmarks/`. Each writes
its RDS output to `inst/extdata/`.

```{r rerun, eval = FALSE}
# Strict-gate null calibration (~5 min, 50 iters x 3 sample sizes)
source("data-raw/benchmarks/bench_null_hires.R")

# Relaxed-gate null calibration
source("data-raw/benchmarks/bench_null_hires_relaxed.R")

# Power benchmark (~30 min, 4 effect sizes x 3 sample sizes x 30 iters)
source("data-raw/benchmarks/bench_power.R")

# Relaxed-gate power
source("data-raw/benchmarks/bench_power_relaxed.R")

# Runtime / memory (~10 min, 3 sizes x 3 reps)
source("data-raw/benchmarks/bench_runtime.R")
```

## Session info

```{r session-info}
sessionInfo()
```
