---
title: "RIPPLE benchmark: FDR calibration, power, and method comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE benchmark: FDR calibration, power, and method comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
library(data.table)
library(ggplot2)
library(patchwork)
```

## Overview

This vignette presents three benchmarks that validate RIPPLE's statistical
properties on synthetic data:

1. **FDR calibration under the null** — empirical type-I error control.
2. **Power curves** — sensitivity as a function of effect size and sample count.
3. **Method comparison** — RIPPLE vs. MISTy and nnSVG on shared synthetic data.

All benchmarks use a common synthetic data generator (`data-raw/benchmarks/`)
that produces `SpatialExperiment` objects with controllable spatial structure
and optional planted distance-dependent gradients.

## Synthetic data design

Each dataset contains three cell types:

- **Tumour** (30 cells): dense central cluster (radius 60 µm).
- **T cells** (120 cells): uniformly distributed; these are the target for
  gradient testing.
- **Fibroblasts** (50 cells): annular ring at 100--200 µm from the tumour
  centre.

For **null** datasets, all genes follow a flat Poisson rate with no distance
dependence. For **gradient** datasets, a subset of genes in T cells have
expression rates modulated by `exp(beta * distance_to_tumour)`, where `beta`
controls the strength and direction of the gradient.

## 1. FDR calibration under the null

50 null datasets (no planted gradient, 50 background genes each) were generated
for each of three sample sizes (N = 3, 5, 10 biological replicates). RIPPLE
was run on each; the empirical FDR is the fraction of genes called significant
at Fisher FDR < 0.05.

```{r fdr-calibration, fig.cap = "Empirical FDR under a true null across 50 iterations per sample size. Dashed line: nominal 0.05 level."}
null_path <- system.file("extdata/bench_null_results.rds", package = "ripple")
if (file.exists(null_path)) {
  null_res <- readRDS(null_path)$per_run

  p1 <- ggplot(null_res, aes(x = factor(n_samples), y = empirical_fdr)) +
    geom_boxplot(fill = "#E8E8E8", outlier.size = 1) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "#D55E00") +
    labs(
      x = "Number of biological replicates (N)",
      y = "Empirical FDR",
      title = "FDR calibration under the null"
    ) +
    theme_minimal()
  print(p1)

  # Summary table
  summary_dt <- null_res[, .(
    runs = .N,
    mean_fdr = sprintf("%.4f", mean(empirical_fdr)),
    max_fdr = sprintf("%.4f", max(empirical_fdr)),
    pooled_fdr = sprintf("%.4f", sum(n_sig_fdr) / sum(n_genes_tested))
  ), by = n_samples]
  knitr::kable(summary_dt,
    col.names = c(
      "N", "Runs", "Mean FDR",
      "Max FDR", "Pooled FDR"
    ),
    caption = "Summary of null FDR calibration."
  )
} else {
  cat("Null benchmark results not yet available.\n")
  cat("Run: Rscript data-raw/benchmarks/bench_null.R\n")
}
```

## 2. Power curves

5 gradient genes (negative coefficient, induced near tumour) and 45 background
genes were planted in each dataset. Four effect sizes (`|beta|` = 0.002, 0.005,
0.01, 0.02) were tested across three sample sizes (N = 3, 5, 10), with 30
iterations each. Power is the fraction of the 5 gradient genes recovered at
FDR < 0.05.

```{r power-curves, fig.cap = "Power to detect planted gradient genes as a function of effect size and sample count. Error bars: ± 1 SD across iterations."}
power_path <- system.file("extdata/bench_power_results.rds", package = "ripple")
if (file.exists(power_path)) {
  power_res <- readRDS(power_path)$summary

  p2 <- ggplot(
    power_res,
    aes(
      x = abs(beta), y = mean_power,
      color = factor(n_samples), group = factor(n_samples)
    )
  ) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    geom_errorbar(
      aes(
        ymin = pmax(0, mean_power - sd_power),
        ymax = pmin(1, mean_power + sd_power)
      ),
      width = 0.001
    ) +
    scale_x_log10(labels = scales::label_number()) +
    labs(
      x = expression("|" * beta * "| (log-rate change per µm)"),
      y = "Power (sensitivity)",
      color = "N (replicates)",
      title = "Power to detect gradient genes"
    ) +
    theme_minimal()
  print(p2)

  knitr::kable(
    power_res[, .(
      beta = sprintf("%.3f", beta), N = n_samples,
      power = sprintf("%.2f", mean_power),
      sd = sprintf("%.2f", sd_power),
      FDR = sprintf("%.3f", mean_fdr)
    )],
    col.names = c("beta", "N", "Power", "SD", "Observed FDR"),
    caption = "Power and observed FDR by effect size and sample count."
  )
} else {
  cat("Power benchmark results not yet available.\n")
  cat("Run: Rscript data-raw/benchmarks/bench_power.R\n")
}
```

## 3. Method comparison (MISTy and nnSVG)

<!-- TODO: Add once bench_comparison.R results are available -->

To be completed: comparison of RIPPLE, MISTy, and nnSVG on the same synthetic
datasets. Expected findings:

- **nnSVG** detects planted genes (positive control) but also flags many
  additional SVGs unrelated to the tumour anchor (lower specificity).
- **MISTy** importance scores correlate with RIPPLE |beta| (agreement on
  signal magnitude) but do not provide direction or calibrated FDR.
- **RIPPLE** uniquely provides directional inference, sign-consistency, and
  formally calibrated FDR.

## Session info

```{r session-info}
sessionInfo()
```
