---
title: "RIPPLE applied to CosMx NSCLC: reproducing paper Figure 4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE applied to CosMx NSCLC: reproducing paper Figure 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
library(ggrepel)
library(patchwork)
```

## Overview

This vignette walks through RIPPLE on the **CosMx SMI NSCLC dataset**
from [He et al. (2022)](https://doi.org/10.1038/s41587-022-01483-z).
765,771 cells, 980 genes, 17 cell types, 5 patients. Each patient is
one biological replicate. Tumor cells are the query. The question:
which genes in the other cell types change expression with distance to
the nearest tumor cell, reproducibly across the 5 patients.

This vignette reproduces panels A to E of **paper Figure 4**. Follow
it to see how to use RIPPLE and what the outputs look like. Each panel
uses cached results bundled with the package. Running the analysis
from scratch on the raw 2.6 GB h5ad takes about 6 hours on a 16-core
machine. We show that step but do not run it.

The required steps are `run_ripple()` and the specificity check.
Everything else is optional. Permutation testing, confounder control,
fGSEA, and ligand-receptor nomination are all downstream stages you
opt into when the question calls for them. Each optional step below
includes a one-line why/when.

> **Note:** for a 1-minute end-to-end on the synthetic mock dataset
> (no download needed), see
> `vignette("getting_started", package = "ripple")` first.

## Dataset preparation (skipped here)

Building the canonical `SpatialExperiment`:

```{r data-prep, eval = FALSE}
library(zellkonverter)
library(SpatialExperiment)

sce <- readH5AD("cosmx_nsclc.h5ad")

# Merge per-patient tumor subtype labels into a single "tumor" class
ct <- as.character(colData(sce)$cell_type)
ct[grepl("^tumor", ct)] <- "tumor"

spe <- SpatialExperiment(
  assays        = list(counts = assay(sce, "counts")),
  colData       = DataFrame(
    cell_type = ct,
    patient   = colData(sce)$patient,
    sample    = colData(sce)$sample
  ),
  spatialCoords = as.matrix(reducedDim(sce, "spatial"))
)
```

## Choosing `k_neighbors` with `plot_k_diagnostics()`

For dense query populations like this one (~150 K tumor cells across 5
patients), `k = 1` is fine. The mean per-cell distance to the nearest
tumor cell stabilises immediately. For sparse query populations (a few
hundred cells per sample), larger `k` (3 to 5) reduces noise.
`plot_k_diagnostics()` gives a quick visual answer:

```{r k-diagnostics, eval = FALSE}
plot_k_diagnostics(
  input           = spe,
  query_celltype  = "tumor",
  celltype_column = "cell_type",
  sample_column   = "patient",
  k_range         = 1:10,
  max_distance_um = 200
)
```

If no clear elbow appears, the query is too sparse for robust per-cell
distance estimation; consider a coarser annotation or more replicates.

## Running RIPPLE

```{r run-ripple, eval = FALSE}
results <- run_ripple(
  input                = spe,
  query_celltype       = "tumor",
  celltype_column      = "cell_type",
  sample_column        = "patient",
  output_dir           = "ripple_output",
  analysis_name        = "tumor_ripple",
  k_neighbors          = 1,
  max_distance_um      = 200,
  min_cells_per_sample = 30,
  n_permutations       = 0,
  verbose              = TRUE
)
```

Key parameter choices:

- **`k_neighbors = 1`**: distance to the single nearest tumor cell;
  appropriate for paracrine signalling range.
- **`max_distance_um = 200`**: cells beyond 200 µm are excluded from the
  GLM. Captures typical secreted-factor gradients without long-range
  dilution.
- **`sample_column = "patient"`**: each patient is one biological
  replicate (N = 5). Per-patient coefficients are combined via Fisher's
  combined p-value with sign-consistency gating.
- **`n_permutations = 0`**: permutation testing skipped here. It is an
  optional validation step (why/when: run it to confirm a gradient is
  specific to the query labels and not an artefact of cell density);
  add it later via `run_permutation_tests()` or the GPU script in
  `inst/python/`.

## Loading cached results

The pre-computed outputs are bundled. In your own analysis, replace
these paths with files under your `output_dir/<analysis_name>/`.

```{r load-results}
results_path <- system.file("extdata/cosmx_cached/all_genes_results.csv",
                            package = "ripple")
results <- fread(results_path)

cat(nrow(results), "gene x cell_type combinations tested\n")
cat(sum(results$fisher_fdr < 0.05, na.rm = TRUE), "significant at Fisher FDR < 0.05\n")
```

## Pre-flight: specificity check (run this early!)

> **Strongly recommended:** run the specificity check **before** any
> downstream interpretation, top-hit listing, or pathway enrichment.
> The **broad-expression class** is the set of genes significant in
> many cell types at once. Common offenders: ribosomal genes, MHC
> housekeeping, ambient RNA from highly expressed query markers. If
> you do not filter them, they dominate every "top hits" table and
> every fGSEA leading edge. The cell-type-specific biology gets
> masked.
>
> **Contamination is user-defined. RIPPLE only proposes candidates.**
> The pipeline gives you three heuristic flags:
>
> 1. The cross-cell-type `broad` class
>    (`classify_gene_specificity()`).
> 2. The ambient-family blocklist (`find_ambient_family_genes()` for
>    Ig, J-chain, Rp, Mt genes).
> 3. A user-supplied query-marker list (`query_signature_genes` on
>    `run_ripple()`).
>
> None of these is a measurement of contamination. A broadly-
> significant gene can be genuine shared biology. Cytokines, MHC II,
> and stress programs are legitimately multi-cell-type. A query
> marker that spills into one neighbouring cell type can be classed
> as `specific` (not `broad`) but still be contamination. See the
> EPCAM-in-fibroblasts case below. You decide what counts as
> contamination using domain knowledge. The flags are starting
> points for inspection, not the truth.

`classify_gene_specificity()` returns one row per significant gene
with two key fields. `n_celltypes` is the number of cell types it
appears in. `specificity_class` is a four-level label: `specific`,
`moderate`, `ubiquitous`, `broad`. We compute it **once**, here, and
reuse it everywhere downstream.

**A note on the broad-class dials.** There is no universally correct
cutoff. `classify_gene_specificity()` exposes **two** dials, both of
which should be reported in your methods:

1. **`broad_threshold`** (how many cell types). Default `4` is
   sensible for the ~10-20 cell-type panel here (and for the original
   HyMy / TDLN analyses), but tune to your data:
   - **Few cell types (≤ 5):** a cutoff of 4 is barely reachable;
     consider 2 or 3.
   - **Many fine subtypes (~30+):** a cutoff of 4 is too lenient and
     will flag legitimate lineage-shared biology; consider 6-10.
   - Sanity check: plot `n_celltypes` as a histogram across all
     significant genes. The cutoff should sit in the right tail,
     separating clear multi-cell-type genes from cell-type-specific
     signal.
2. **`broad_sig_threshold`** (how strictly significant each "hit"
   must be). Default `NULL` means any `*`-significant (FDR < 0.05)
   hit counts toward the broad-class tally. Tighten to `"**"`
   (FDR < 0.01) or `"***"` (FDR < 0.001) to make the filter **looser
   overall**: a gene only counts as a hit in cell types where the
   signal is unambiguous, so fewer genes get flagged. Useful when the
   default classifies as "broad" things you believe are real biology.

```{r preflight-broad}
spec <- classify_gene_specificity(
  results,
  fdr_threshold       = 0.05,
  broad_threshold     = 4,     # tune for your panel; see note above
  broad_sig_threshold = NULL   # NULL == any "*" hit counts;
                                # use "**" / "***" to require
                                # stricter sig per hit
)

table(spec$specificity_class)

# Genes flagged as broad, to be deprioritised in top-hit lists
broad_genes <- spec[specificity_class == "broad", unique(gene)]
cat("Broad-class genes:", length(broad_genes), "\n")
head(spec[specificity_class == "broad"][order(-n_celltypes)], 10)
```

```{r preflight-clean}
# Filtered copy used downstream when picking labels / top-hit lists.
# Plots that show *all* genes as background (volcanos, fGSEA) still use
# `results`; only the curated label / top-hit selections come from
# `results_clean`. This way a flagged gene like MIF is still visible as
# a dot on the volcano, but it does NOT crowd out the cell-type-specific
# hits in the labelled set.
results_clean <- results[!gene %in% broad_genes]
cat(nrow(results),       "rows in `results` (unfiltered)\n")
cat(nrow(results_clean), "rows in `results_clean` (broad class removed)\n")
```

> **⚠ Limitations of broad-class filtering (read before reporting
> filtered results):**
>
> - **Loses real biology when broad-expression and induction
>   co-occur.** A gene can be both genuinely induced near the query
>   *and* inflated by ambient RNA from query cells. MIF in this
>   dataset is a textbook example. Filtering removes the gene from
>   every cell type's ranking. The cell types where the induction is
>   real lose it too. Always inspect the broad-class list and decide
>   case-by-case.
> - **The threshold is panel-dependent.** Default `broad_threshold = 4`
>   assumes 10 to 20 cell types in the panel. With 3 cell types the
>   cutoff is unreachable. With 30 fine subtypes it is too lenient.
>   Re-tune for your annotation granularity.
> - **Removes genes globally, not per cell type.** A gene flagged
>   dataset-wide is dropped from every column. A gene that is
>   broad-expression across most cell types may still be a real
>   signal in the one cell type that biologically expresses it.
> - **Multi-cell-type biology is not ambient RNA.** Cytokines, MHC
>   class II genes, and stress-response programs are legitimately
>   expressed across many cell types. They land in the broad class.
>   The heuristic cannot tell them apart from ambient RNA on its own.
> - **Rank-based tests are panel-size sensitive.** Dropping genes
>   shifts the ranks of all remaining genes. NES values from
>   filtered and unfiltered fGSEA runs are not directly comparable.
>   Always report whether the filter was applied. Ideally show both.
>
> **Recommended workflow:**
>
> 1. Classify upfront (this section).
> 2. Keep both `results` (full) and `results_clean` (filtered) in
>    scope.
> 3. Plot all genes as background. Label and report top hits from
>    `results_clean`.
> 4. Run fGSEA both with and without `exclude_broad = TRUE` (Panel D
>    below).

# Reproducing paper Figure 4

The full multi-panel figure is rendered by `paper/render_figures.R`
(saves to `paper/figures/fig_4_cosmx.{pdf,png}`). Here we walk through
each panel with the same cached inputs.

## Panel A: gradients per target cell type

```{r fig4-A, fig.width = 9, fig.height = 5}
plot_gene_counts_by_celltype(
  results,
  query_label     = "Tumor",
  broad_threshold = 4
)
```

Each bar is split two ways. By direction: induced near tumor vs
repressed near tumor. By specificity class: solid for cell-type
specific, faded for the "ubiquitous / broad" fraction (genes
significant in >= 4 cell types). This is the key QC view. A tall bar
only matters biologically if its solid portion dominates. Fibroblasts
and macrophages are the most remodelled cell types near tumor. Both
show a sizeable specific-to-broad ratio.

## Panel B: fibroblast gradient volcano

The default `plot_gradient_volcano()` works well as a quick look:

```{r fig4-B-quick, fig.width = 8, fig.height = 6}
plot_gradient_volcano(
  results[cell_type == "fibroblast"],
  query_label = "Tumor",
  n_label     = 15
)
```

For the published figure we colour by direction (red = induced, blue
= repressed) and label a curated gene set. The labelling strategy is
worth noting. **All** genes still appear as background dots, so the
broad-class genes remain visible for interpretation. But the labels
go on genes that are NOT in the broad list. We add a few flagged
genes by hand (MIF, IGKC) so we can discuss them.

```{r fig4-B-paper, fig.width = 8, fig.height = 6}
fib <- results[cell_type == "fibroblast"]    # all genes for the scatter
fib[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
fib[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Cell-type-specific top hits, pulled from `results_clean` so that
# ambient RNA does not crowd out the cell-type-specific biology.
top_clean_fib <- results_clean[
  cell_type == "fibroblast" & fisher_fdr < 0.05
][order(fisher_fdr)][1:12]$gene

# Plus a few flagged genes we want to discuss explicitly (see "Key
# fibroblast hits" below). MIF, IGKC and RPL22 are broad-class
# (significant in >= 4 cell types); JCHAIN is NOT broad (it is
# significant only in fibroblasts here) but is an ambient-family gene
# caught by find_ambient_family_genes() instead. Two different flags.
discuss_flagged <- c("MIF", "IGKC", "JCHAIN", "RPL22")

label_fib <- unique(c(top_clean_fib, discuss_flagged))
fib[, to_label := gene %in% label_fib & fisher_fdr < 0.05]

ggplot(fib, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = fib[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = 20, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = "Gradient score (per um)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Fibroblasts: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

**Key fibroblast hits (from `results_clean`):**

- **IGFBP5** (induced): IGF-binding protein involved in stromal
  remodelling; consistent with cancer-associated fibroblast (CAF)
  activation.
- **CXCL12** (repressed): chemokine normally expressed by quiescent
  fibroblasts; its repression near tumor reflects phenotypic switching
  to a CAF state.

**Flagged hits (kept on the plot, excluded from the clean
top-hit list):**

- **MIF** (induced): macrophage migration inhibitory factor, a
  well-known tumor-derived cytokine. *However*, MIF also appears as
  significant in 9 cell types and is flagged by the broad-class
  heuristic. This is the canonical "broad expression + induction"
  case from the limitations note above: MIF is plausibly both a real
  fibroblast signal AND inflated by ambient RNA from highly
  MIF-expressing tumor cells. We still show it on the plot, but we
  did not let it compete for label slots with cell-type-specific
  hits.
- **IGKC** (broad, significant in 4 cell types) and **RPL22** (broad,
  8 cell types): an immunoglobulin and a ribosomal gene flagged by the
  cross-cell-type heuristic, almost certainly ambient RNA / leakage.
- **JCHAIN** (induced): a plasma-cell J-chain gene that is significant
  only in fibroblasts here, so it is `specific`, NOT `broad`. It is
  still suspicious ambient RNA, and `find_ambient_family_genes()`
  catches it via the Ig/J-chain family pattern. This is exactly why the
  flags are complementary: the cross-cell-type `broad` flag misses
  focal spillover, but the ambient-family list catches it.

## Panel C: per-patient consistency for MIF

A key feature of RIPPLE is per-replicate fitting. The forest plot below
shows per-patient gradient scores ± 95 % CI for MIF in fibroblasts:

```{r fig4-C, fig.width = 7, fig.height = 3.5}
coef_path <- system.file(
  "extdata/cosmx_cached/per_celltype/fibroblast/coef_per_sample.csv",
  package = "ripple"
)
coef_data <- fread(coef_path)

mif <- coef_data[gene == "MIF"]
mif[, ci_lo := coef - 1.96 * se]
mif[, ci_hi := coef + 1.96 * se]

ggplot(mif, aes(x = coef, y = sample_id, colour = sample_id)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi),
                orientation = "y", width = 0.25, alpha = 0.7) +
  geom_point(size = 3) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(
    x = "Per-patient gradient score for MIF (per um)",
    y = NULL,
    title = "MIF in fibroblasts: per-patient gradient scores"
  ) +
  theme_ripple(base_size = 12)
```

All five patients show a consistent negative coefficient. MIF RNA
counts increase near tumor in every replicate. Reproducibility alone
does not rule out systematic ambient RNA. Both real induction and
ambient RNA can be reproducible across replicates. That is why the
specificity check is a separate QC step.

## Panel D: pathway enrichment (fGSEA dotplot)

fGSEA is an **optional** step (why/when: run it to summarise dozens of
significant genes per cell type into interpretable pathway-level
programs; skip it if you only care about individual gene hits).

```{r fig4-D-run, eval = FALSE}
# Run fGSEA via the atlas wrapper (writes fgsea_all_celltypes.csv)
run_ripple_atlas(
  results_dir = "ripple_output/tumor_ripple",
  query_label = "Tumor",
  run_fgsea   = TRUE,
  organism    = "human",
  fgsea_seed  = 42
)
```

```{r fig4-D-load}
fgsea_path <- system.file("extdata/cosmx_cached/fgsea_all_celltypes.csv",
                          package = "ripple")
fgsea_dt <- fread(fgsea_path)
```

`plot_fgsea_dotplot()` summarises the table at a glance: dot size
encodes -log10(padj), dot colour encodes NES on the same red/blue
diverging scale as the gradient volcano (red NES = induced near tumor,
blue NES = repressed), and non-significant cells are dimmed.

```{r fig4-D, fig.width = 9, fig.height = 7}
plot_fgsea_dotplot(fgsea_dt, padj_threshold = 0.1, top_n = 12)
```

Hallmark pathways like **E2F_TARGETS**, **G2M_CHECKPOINT** (proliferation,
red is induced in tumor-proximal fibroblasts and macrophages) and
**ALLOGRAFT_REJECTION** (immune effector programs, blue is depleted in
tumor-proximal myeloid + lymphoid cells) light up across multiple
stromal/immune cell types.

### Robust pathway enrichment: dropping broad-class genes

The broad-class genes also dominate fGSEA leading edges. Ribosomal
genes, MHC housekeeping, and ambient-RNA-prone markers light up
`INTERFERON_GAMMA_RESPONSE`, `TNFA_SIGNALING_VIA_NFKB`, and other
pathways in cell types where they do not really belong.
`run_ripple_fgsea()` accepts `exclude_broad = TRUE` to drop them
before ranking. It uses the same heuristic as the upfront
classification above.

A second clean-up is to remove a **user-defined query signature**
before ranking. The cross-cell-type `broad` flag catches genes that
light up in many cell types. It does not catch *focal* spillover. A
tumor marker that bleeds into one neighbouring cell type scores as
`specific`, not `broad`. Examples in this dataset: EPCAM and the
keratins, which show up in fibroblasts. Define your query-signature
list once from domain knowledge. Reuse it for every downstream step
(fGSEA, L-R, anything else).

```{r fgsea-filtered, eval = FALSE}
# Query (tumor/epithelial) signature: focal spillover the broad flag misses.
# Define from domain knowledge and reuse across all downstream steps.
tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "KRT7", "KRT17")

# Using the same threshold we picked in the pre-flight section.
fgsea_filtered <- run_ripple_fgsea(
  results,
  gene_sets             = "hallmark",
  organism              = "human",
  exclude_broad         = TRUE,
  broad_threshold       = 4,
  query_signature_genes = tumor_markers   # drop tumor spillover before ranking
)

# Side-by-side: with and without filter
plot_fgsea_dotplot(fgsea_dt,       padj_threshold = 0.1, top_n = 12) +
  ggplot2::labs(title = "fGSEA -- all genes")
plot_fgsea_dotplot(fgsea_filtered, padj_threshold = 0.1, top_n = 12) +
  ggplot2::labs(title = "fGSEA -- broad-class genes excluded")
```

If the two panels look similar, the unfiltered enrichment is stable
against the most obvious ambient RNA. If they differ a lot, the
filter is removing real signal, broad-class biology, or both.
Investigate the dropped leading-edge genes before reporting either
version. The full caveat list is in the
[Pre-flight: specificity check] section above and in
`?run_ripple_fgsea`. The most important one for fGSEA: NES values are
panel-size-sensitive. They are **not directly comparable** between
filtered and unfiltered runs. (`run_ripple_fgsea()` also accepts
`exclude_specificity_class = "broad"` as an equivalent way to drop the
broad class.)

## Panel E: top tumor → fibroblast ligand-receptor pairs

Ligand-receptor nomination is an **optional** step (why/when: run it to
turn a list of gradient genes into testable signalling hypotheses by
matching them to known L-R pairs; skip it if you do not need a
mechanistic, pair-level read-out).

Stage 6 (`run_ripple_lr()`) intersects gradient genes with curated
ligand-receptor databases and (optionally) NicheNet, producing a
combined prioritization score per L-R pair.

**Directional logic (why receptors must be induced).** Signalling is
nominated query → receiver. The tumor sends a ligand. The neighbouring
fibroblast carries the receptor. For that to be coherent, the
receptor RNA should go UP as cells get closer to the tumor (induced,
negative gradient score). A receptor that is *repressed* near the
tumor is not a plausible "tumor signals to fibroblast" axis. So
`run_ripple_lr()` nominates induced receptors only by default:

```{r fig4-E-run, eval = FALSE}
lr_out <- run_ripple_lr(
  results_dir           = "ripple_output/tumor_ripple",
  input                 = spe,
  query_celltype        = "tumor",
  celltype_column       = "cell_type",
  sample_column         = "patient",
  organism              = "human",
  receptor_direction    = "induced",  # default: only receptors UP near tumor
  exclude_contamination = TRUE,       # default: drop cross-cell-type broad receptors
  broad_threshold       = 4L,         # same broad cutoff as the pre-flight
  query_signature_genes = tumor_markers  # drop tumor-marker receptors (e.g. EPCAM)
)
```

`exclude_contamination = TRUE` (the default) also drops receptors
flagged as cross-cell-type `broad`. The nomination is not driven by
ambient RNA. Set `receptor_direction = "repressed"` or `"both"` only
if you have a specific reason to look at the other directions.

> **Known limitation: focal query-marker spillover is not caught by
> the cross-cell-type flag.** EPCAM is a tumor/epithelial marker. In
> this dataset it scores as `induced` and `specific` (significant
> only in fibroblasts). So `receptor_direction = "induced"` does not
> remove it, and the cross-cell-type `broad` filter does not remove
> it either. CD44 → EPCAM lands at the top of the list. It is almost
> certainly EPCAM RNA bleeding from tumor cells into fibroblast
> segments at the interface, not a real fibroblast receptor. The
> cross-cell-type flag cannot catch it because the spillover is
> focal (one cell type only). That is what `query_signature_genes`
> is for. Passing the tumor markers (including EPCAM) drops them as
> receptors. CD44 → EPCAM disappears and collagen → DDR1 becomes the
> top pair. Automated flags are starting points. The query
> signature is yours to define.

The bundled `combined_prioritization.csv` is the output of the
cleaned `run_ripple_lr()`. Induced receptors only, with cross-cell-
type broad receptors and the tumor/epithelial query signature already
excluded. Every pair has a negative (induced) receptor gradient
score. The top pair is COL1A1 → DDR1 instead of the CD44 → EPCAM
spillover artefact:

```{r fig4-E-load}
lr_path <- system.file(
  "extdata/cosmx_cached/lr/fibroblast/combined_prioritization.csv",
  package = "ripple"
)
lr_fib <- fread(lr_path)

cat(nrow(lr_fib), "tumor -> fibroblast L-R pairs (all induced receptors)\n")
head(lr_fib[order(-combined_score),
            .(ligand, receptor, combined_score, receptor_gradient_coef,
              receptor_fdr, ligand_pct_query)], 5)
```

```{r fig4-E, fig.width = 9, fig.height = 6}
lr_top <- lr_fib[order(-combined_score)][1:15]
lr_top[, pair := factor(paste0(ligand, " -> ", receptor),
                        levels = rev(paste0(ligand, " -> ", receptor)))]

ggplot(lr_top, aes(x = combined_score, y = pair)) +
  geom_point(aes(size = ligand_pct_query), colour = "#B2182B") +
  scale_size_continuous(range = c(2, 7),
                        name = "Tumor cells\nexpressing ligand (%)") +
  labs(
    x = "Combined prioritization score",
    y = NULL,
    title = "Top tumor -> fibroblast L-R pairs"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "right",
        legend.box = "vertical",
        legend.key.size = unit(0.4, "cm"))
```

After CD44 → EPCAM (the spillover artefact flagged above), the top
induced pairs are ECM-to-receptor axes: collagens and fibronectin
signalling to **DDR1** and **ITGB8** (COL1A1 → DDR1, FN1 → ITGB8,
COL1A2 → DDR1). These match the ECM-remodelling hits in Panel B and
the matrix/EMT pathway enrichment in Panel D. Gradient, pathway, and
L-R views all converge on the same stromal-remodelling biology. CD44
→ EPCAM at the top is the cautionary counter-example: a high score
that is an artefact, not biology. That is why the query-signature
check matters.

## Reproducing the full multi-panel figure

The exact PDF used in the paper is rendered by:

```{r render-fig4, eval = FALSE}
source("paper/render_figures.R")
# writes paper/figures/fig_4_cosmx.{pdf,png}
```

# Beyond Figure 4: macrophage deep-dive, specificity, and QC

## Macrophage gradient volcano

Macrophages show a polarised response near tumor (94 significant
genes). Induced programs: motility, interferon, plus some tumor-RNA
spillover. Repressed programs: antigen-presentation and homeostatic
/ complement. The same volcano view as Panel B, applied to
macrophages:

```{r mac-volcano, fig.width = 8, fig.height = 6}
mac <- results[cell_type == "macrophage"]   # all genes for the scatter
mac[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
mac[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Same labelling pattern as the fibroblast panel:
# top hits from `results_clean` (cell-type-specific signal) plus a
# curated set of biologically interesting macrophage markers (some of
# which, e.g. MHC II, are multi-cell-type by nature and therefore
# appear in `broad_genes`; we keep them because the macrophage gradient
# is the biological point of interest). Only genes with fisher_fdr <
# 0.05 in macrophages are actually labelled (the `to_label` gate below).
top_clean_mac <- results_clean[
  cell_type == "macrophage" & fisher_fdr < 0.05
][order(fisher_fdr)][1:10]$gene

curated_mac <- c("HLA-DRA", "HLA-DPA1", "HLA-DRB1", "CD74", "HLA-E",
                 "C1QA", "SELENOP", "CD68", "MS4A4A",
                 "S100A6", "MIF", "CXCL10", "TUBB", "SPP1")

label_mac <- unique(c(top_clean_mac, curated_mac))
mac[, to_label := gene %in% label_mac & fisher_fdr < 0.05]

ggplot(mac, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = mac[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = 25, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = "Gradient score (per um)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Macrophages: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

Read directions off the cached coefficients, not off expectations.
The macrophage axis here runs the opposite way to the textbook TAM
cartoon. That is itself informative.

**Macrophage induced near tumor (negative gradient score):**

- **S100A6, TUBB** (cytoskeletal / calcium-binding): motility and
  morphological adaptation in the tumor niche.
- **CXCL10**: an interferon-induced chemokine, consistent with an
  inflamed tumor-proximal niche.
- **MIF** (also broad-class): tumor paracrine signal plus likely
  ambient RNA, same caveat as in fibroblasts.
- **KRT19, SPP1** (epithelial / tumor markers): induced and `specific`
  to macrophages, but very likely tumor-RNA spillover into macrophage
  segments at the interface, NOT macrophage biology. This is the same
  query-marker spillover pattern as EPCAM in fibroblasts; a
  query-signature check would flag it.

**Macrophage repressed near tumor (positive gradient score):**

- **HLA class II** (HLA-DRA, HLA-DPA1, HLA-DRB1, HLA-E) and **CD74**:
  antigen-presentation machinery is *down* near tumor, consistent with
  reduced MHC II on tumor-proximal macrophages rather than the
  intratumoral upregulation one might assume.
- **SELENOP, C1QA, CD68, MS4A4A**: complement and homeostatic /
  tissue-resident macrophage programs, also lower near tumor.

Both the antigen-presentation axis and the homeostatic / complement
axis are repressed near tumor. Motility, interferon, and tumor
spillover signals come up. The polarisation read-out is data-driven
here. It is not a re-statement of a prior expectation.

# Specificity breakdown plot and QC dashboard

The specificity check was already run upfront (see
[Pre-flight: specificity check]). The `spec` object and `broad_genes`
vector are still in scope. `plot_specificity_breakdown()` summarises
the four classes as a one-glance bar:

```{r breakdown, fig.width = 7, fig.height = 4}
plot_specificity_breakdown(results, broad_threshold = 4)
```

Genes like **HLA-E**, **KRT19**, **MIF**, and **RPL22** are flagged
because they are significant across 8 to 9 cell types. That is a
hallmark of ambient RNA. MIF (significant in 9 cell types) is the
canonical example of why filtering must be deliberate. MIF is
plausibly both genuinely induced by tumor paracrine signalling **and**
inflated by ambient RNA from high-expressing tumor cells.

## QC dashboard

The CosMx run was launched without a per-cell distance dump (older
version). `ripple_plot_qc()` here will skip panels 1+2 and warn. That
is expected. On any fresh `run_ripple()` call all seven panels are
populated automatically.

```{r qc, eval = FALSE}
qc <- ripple_plot_qc(
  results_dir           = "ripple_output/tumor_ripple",
  query_signature_genes = c("KRT19", "EPCAM", "MKI67", "KRT8"),
  query_label           = "Tumor"
)
```

The bleed-through panel (panel 7) takes a query-marker vector. Any
marker that shows up significant in many target cell types is
highlighted in red. That is the smoking gun for ambient RNA.

# Interpretation summary

This CosMx NSCLC analysis demonstrates RIPPLE's ability to detect
biologically interpretable spatial gradients in a multi-patient dataset:

1. **Fibroblasts** show the strongest tumor-proximity remodelling
   (129 genes). IGFBP5 is induced near tumor (a CAF-activation
   marker). CXCL12, THBS1, and MMP2 are repressed. Top hits flagged
   in the broad-expression class (MIF, KRT19) should be validated
   independently.
2. **Macrophages** show a polarised response (94 genes). Near tumor:
   motility / calcium-binding (S100A6, TUBB) and interferon (CXCL10)
   genes are induced. Antigen-presentation (HLA-DRA, HLA-DPA1, CD74)
   and homeostatic / complement (SELENOP, C1QA) programs are
   repressed. KRT19 and SPP1 also score as induced. They are likely
   tumor-RNA spillover, not macrophage biology.
3. **T cell subsets** each show distinct gradient signatures (CD4
   memory: 55 genes, CD8 memory: 27, Tregs: 23). The adaptive immune
   response is spatially organised around tumor nests.
4. **Hallmark pathway enrichment** (EMT, proliferation, immune
   effector programs) recovers known tumor-niche biology. It ties
   stromal and immune programs together.
5. **L-R prioritization** (induced receptors only) nominates
   ECM-to-receptor axes (COL1A1 → DDR1, FN1 → ITGB8) as the top
   tumor → fibroblast candidates. The gradient and pathway panels
   corroborate them. The highest-scoring pair, CD44 → EPCAM, is an
   EPCAM spillover artefact. A high score is a hypothesis to check,
   not a conclusion.
6. **Contamination flagging** proposes candidates through three
   complementary heuristics: cross-cell-type broad, ambient-family
   blocklist, query-marker spillover. They are starting points for
   the user to curate with domain knowledge. Not a definitive call.

All 516 significant genes are reproduced across 5 patients with
unanimous sign consistency (the strict gate). Strong evidence for
genuine tumor-microenvironment spatial interactions.

## Session info

```{r session-info}
sessionInfo()
```
