---
title: "RIPPLE applied to CosMx NSCLC: Tumor-proximity gene expression gradients"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE applied to CosMx NSCLC: Tumor-proximity gene expression gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
library(ggrepel)
```

## Overview

This vignette demonstrates RIPPLE on a real spatial transcriptomics dataset:
the **CosMx SMI NSCLC dataset** from [He et al. (2022)](https://doi.org/10.1038/s41587-022-01483-z),
comprising 765,771 cells across 5 patients with non-small cell lung cancer.

We use **tumor cells** as the query population and ask: *which genes in the
surrounding cell types change expression as a function of physical distance
to the nearest tumor cell?*

The analysis reveals 516 significant distance-dependent gene gradients across
17 target cell types, with fibroblasts (129 genes) and macrophages (94 genes)
showing the strongest tumor-proximity transcriptional remodeling.

## Dataset preparation

The CosMx NSCLC data is publicly available as an h5ad file. The preparation
involves loading the data, merging patient-specific tumor subtype labels into
a single "tumor" class, and building a `SpatialExperiment` object. This step
is shown but not executed here (the raw data is 2.6 GB).

```{r data-prep, eval = FALSE}
library(zellkonverter)
library(SpatialExperiment)

sce <- readH5AD("cosmx_nsclc.h5ad")

# Merge per-patient tumor subtype labels
ct <- as.character(colData(sce)$cell_type)
ct[grepl("^tumor", ct)] <- "tumor"

# Build a SpatialExperiment with clean metadata
spe <- SpatialExperiment(
  assays = list(counts = assay(sce, "counts")),
  colData = DataFrame(
    cell_type = ct,
    patient   = colData(sce)$patient,
    sample    = colData(sce)$sample
  ),
  spatialCoords = as.matrix(reducedDim(sce, "spatial"))
)
```

The resulting object has `r formatC(765771, big.mark = ",")` cells, 980 genes,
18 cell types (including the merged tumor class), and 5 biological replicates
(patients).

## Running RIPPLE

```{r run-ripple, eval = FALSE}
library(ripple)

results <- run_ripple(
  input                = spe,
  query_celltype       = "tumor",
  celltype_column      = "cell_type",
  sample_column        = "patient",
  output_dir           = "ripple_output",
  analysis_name        = "tumor_ripple",
  k_neighbors          = 1,
  max_distance_um      = 200,
  min_cells_per_sample = 30,
  n_permutations       = 0,
  verbose              = TRUE
)
```

Key parameter choices:

- **`k_neighbors = 1`**: distance to the single nearest tumor cell — the
  default for paracrine signaling range.
- **`max_distance_um = 200`**: cells beyond 200 µm are excluded from the
  model. This captures the typical range of secreted factor gradients while
  excluding cells too distant for meaningful interaction.
- **`sample_column = "patient"`**: each patient is one biological replicate
  (N = 5). Per-patient coefficients are combined via Fisher's method with
  sign-consistency gating.
- **`n_permutations = 0`**: permutation testing skipped in this run; can be
  added later with `run_permutation_tests()`.

## Loading cached results

The pre-computed results are bundled with the package. In your own analysis,
replace these paths with your `output_dir`.

```{r load-results}
results_path <- system.file("extdata/cosmx_cached/all_genes_results.csv",
  package = "ripple"
)
results <- fread(results_path)

cat(nrow(results), "gene x cell_type combinations tested\n")
cat(sum(results$fisher_fdr < 0.05, na.rm = TRUE), "significant at FDR < 0.05\n")
```

## Summary of significant hits

```{r sig-summary, fig.height = 5}
sig <- results[fisher_fdr < 0.05]

sig_by_ct <- sig[, .(
  n_induced   = sum(median_coef < 0),
  n_repressed = sum(median_coef > 0)
), by = cell_type]
sig_by_ct[, n_total := n_induced + n_repressed]
sig_by_ct <- sig_by_ct[order(-n_total)]

plot_data <- melt(sig_by_ct,
  id.vars = "cell_type",
  measure.vars = c("n_induced", "n_repressed"),
  variable.name = "direction", value.name = "count"
)
plot_data[, direction := fifelse(
  direction == "n_induced",
  "Induced (near tumor)", "Repressed (near tumor)"
)]
plot_data[, cell_type := factor(cell_type, levels = sig_by_ct$cell_type)]

ggplot(plot_data, aes(x = cell_type, y = count, fill = direction)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Induced (near tumor)" = "#D55E00",
    "Repressed (near tumor)" = "#0072B2"
  )) +
  labs(
    x = NULL, y = "Number of significant genes",
    title = "Tumor-proximity gene gradients per cell type",
    fill = "Direction"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Fibroblasts and macrophages are the most transcriptionally remodeled cell types
near tumor cells, consistent with their established roles in the tumor
microenvironment. Notably, the majority of significant genes in fibroblasts are
**induced** near tumor cells — reflecting activation of cancer-associated
fibroblast (CAF) programs.

## Gradient volcano plot

The gradient volcano shows effect size (median coefficient across patients) vs.
significance (Fisher's combined FDR). Negative coefficients mean gene expression
**increases** near tumor cells (induced); positive coefficients mean it
**decreases** (repressed).

```{r volcano-fibroblast, fig.width = 8, fig.height = 6}
fib <- results[cell_type == "fibroblast"]
fib[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
fib[, significant := fisher_fdr < 0.05]

top_genes <- fib[fisher_fdr < 0.05][order(fisher_fdr)][1:20]

ggplot(fib, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(color = significant), size = 1.2, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("TRUE" = "#D55E00", "FALSE" = "grey70"),
    guide = "none"
  ) +
  ggrepel::geom_text_repel(
    data = top_genes,
    aes(label = gene),
    size = 3, max.overlaps = 20, seed = 42
  ) +
  labs(
    x = "Median coefficient (log-rate change per µm)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Fibroblasts: tumor-proximity gene gradients"
  ) +
  theme_minimal()
```

**Key fibroblast hits:**

- **MIF** (negative coef = induced near tumor): macrophage migration inhibitory
  factor — a known tumor-derived cytokine that activates CAFs.
- **IGFBP5** (negative coef = induced): insulin-like growth factor binding
  protein 5, involved in stromal remodeling.
- **CXCL12** (positive coef = repressed near tumor): a chemokine normally
  expressed by quiescent fibroblasts; its repression near tumor reflects
  phenotypic switching to a CAF state.
- **IGKC, JCHAIN, IGHG2** (positive coef = repressed): immunoglobulin genes
  detected in fibroblasts — these are likely ambient RNA contamination from
  nearby plasma cells (see contamination check below).

## Per-replicate consistency

A key feature of RIPPLE is per-replicate fitting. Here we show the per-patient
coefficients for MIF in fibroblasts:

```{r forest-mif, fig.width = 7, fig.height = 3.5}
coef_path <- system.file(
  "extdata/cosmx_cached/per_celltype/fibroblast/coef_per_sample.csv",
  package = "ripple"
)
coef_data <- fread(coef_path)

mif <- coef_data[gene == "MIF"]
mif[, sample_id := paste0("Patient ", seq_len(.N))]

ggplot(mif, aes(x = coef, y = sample_id)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = coef - 1.96 * se, xmax = coef + 1.96 * se),
    height = 0.2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Coefficient (log-rate change per µm)",
    y = NULL,
    title = "MIF in fibroblasts: per-patient coefficients"
  ) +
  theme_minimal()
```

All five patients show a consistent negative coefficient — MIF expression
increases near tumor cells in every replicate. This reproducibility across
independent biological replicates, enforced by RIPPLE's sign-consistency gate,
gives confidence that MIF induction is a genuine biological signal rather than
a technical artifact.

## Macrophage deep-dive

```{r volcano-macrophage, fig.width = 8, fig.height = 6}
mac <- results[cell_type == "macrophage"]
mac[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
mac[, significant := fisher_fdr < 0.05]

top_mac <- mac[fisher_fdr < 0.05][order(fisher_fdr)][1:15]

ggplot(mac, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(color = significant), size = 1.2, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("TRUE" = "#D55E00", "FALSE" = "grey70"),
    guide = "none"
  ) +
  ggrepel::geom_text_repel(
    data = top_mac,
    aes(label = gene),
    size = 3, max.overlaps = 20, seed = 42
  ) +
  labs(
    x = "Median coefficient (log-rate change per µm)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Macrophages: tumor-proximity gene gradients"
  ) +
  theme_minimal()
```

Macrophages show a mix of induced and repressed programs near tumor cells,
consistent with tumor-associated macrophage (TAM) polarization.

## Contamination check

Genes that are significant in many cell types simultaneously may reflect
ambient RNA contamination or segmentation artifacts rather than true
distance-dependent biology. RIPPLE flags these automatically:

```{r contamination}
contam_path <- system.file("extdata/cosmx_cached/contamination_candidates.csv",
  package = "ripple"
)
contam <- fread(contam_path)

knitr::kable(contam[1:10], caption = "Top contamination candidates")
```

Genes like **IGKC**, **HLA-E**, **KRT19**, and **RPL22** appear significant
across 8-9 cell types — a hallmark of ambient RNA. When interpreting results,
these should be deprioritised or validated independently (e.g., by RNA FISH).

## Pathway enrichment (fGSEA)

RIPPLE integrates with fGSEA to test whether gradient genes are enriched for
known biological pathways:

```{r run-fgsea, eval = FALSE}
run_ripple_atlas(
  results_dir = "ripple_output/tumor_ripple",
  query_label = "Tumor",
  run_fgsea   = TRUE,
  organism    = "human",
  fgsea_seed  = 42
)
```

```{r fgsea-results, fig.width = 9, fig.height = 6}
fgsea_path <- system.file("extdata/cosmx_cached/fgsea_all_celltypes.csv",
  package = "ripple"
)
fgsea_dt <- fread(fgsea_path)

# Top pathways across cell types
sig_pathways <- fgsea_dt[padj < 0.1]
cat(
  length(unique(sig_pathways$pathway_clean)), "pathways enriched (padj < 0.1) across",
  length(unique(sig_pathways$cell_type)), "cell types\n"
)

# Heatmap of NES per cell type for top pathways
if (nrow(sig_pathways) > 0) {
  top_pw <- sig_pathways[, .(mean_abs_nes = mean(abs(NES))), by = pathway_clean]
  top_pw <- top_pw[order(-mean_abs_nes)][1:min(15, .N)]

  plot_dt <- fgsea_dt[pathway_clean %in% top_pw$pathway_clean]
  plot_dt[, pathway_clean := factor(pathway_clean, levels = rev(top_pw$pathway_clean))]
  plot_dt[, sig_label := fifelse(padj < 0.1, "*", "")]

  ggplot(plot_dt, aes(x = cell_type, y = pathway_clean, fill = NES)) +
    geom_tile(color = "white", linewidth = 0.3) +
    geom_text(aes(label = sig_label), size = 4) +
    scale_fill_gradient2(
      low = "#0072B2", mid = "white", high = "#D55E00",
      midpoint = 0, name = "NES"
    ) +
    labs(
      x = NULL, y = NULL,
      title = "fGSEA: MSigDB Hallmark pathway enrichment in tumor-proximity gradients",
      subtitle = "* = padj < 0.1"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  message(
    "No pathways reached padj < 0.1. Consider relaxing the threshold ",
    "or using a different gene set collection."
  )
}
```

## Interpretation summary

This CosMx NSCLC analysis demonstrates RIPPLE's ability to detect biologically
interpretable spatial gradients in a multi-patient dataset:

1. **Fibroblasts** show the strongest tumor-proximity remodeling (129 genes),
   dominated by CAF activation markers (MIF, IGFBP5, THBS1) and ECM genes.
2. **Macrophages** show a polarized response (94 genes), with tumor-proximity
   induction of antigen presentation (CD74, HLA-DRA) and suppression of
   inflammatory programs.
3. **T cell subsets** (CD4 memory: 55 genes, CD8 memory: 27 genes, Tregs: 23
   genes) each show distinct gradient signatures, suggesting spatial
   organisation of the adaptive immune response around tumor nests.
4. **Contamination flagging** correctly identifies ambient RNA artifacts
   (immunoglobulin genes appearing in non-immune cells) — critical for avoiding
   false biological conclusions.

All 516 significant genes are reproduced across 5 independent patients with
unanimous sign consistency, providing strong evidence for genuine
tumor-microenvironment spatial interactions.

## Session info

```{r session-info}
sessionInfo()
```
