---
title: "RIPPLE applied to CosMx NSCLC: reproducing paper Figure 4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE applied to CosMx NSCLC: reproducing paper Figure 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
library(ggrepel)
library(patchwork)
```

## Overview

This vignette walks through RIPPLE on the **CosMx SMI NSCLC dataset**
from [He et al. (2022)](https://doi.org/10.1038/s41587-022-01483-z) —
765,771 cells, 980 genes, 17 cell types, 5 NSCLC patients (each treated
as a biological replicate). Tumor cells are the query; we ask which
genes in the surrounding cell types change expression as a function of
distance to the nearest tumor cell, reproducibly across all 5 patients.

The analysis recovers **516 significant gradient hits** at Fisher
FDR < 0.05 across 17 target cell types, with fibroblasts (129) and
macrophages (94) showing the strongest tumor-proximity remodelling.

This vignette is structured to **directly reproduce panels A–E of
paper Figure 4**. Each panel uses cached results bundled with the
package; running the analysis from scratch on the raw 2.6 GB h5ad takes
~6 hours on a 16-core machine and is shown but not executed.

> **Note:** for a 1-minute end-to-end on the synthetic mock dataset
> (no download needed), see
> `vignette("getting_started", package = "ripple")` first.

## Dataset preparation (skipped here)

Building the canonical `SpatialExperiment`:

```{r data-prep, eval = FALSE}
library(zellkonverter)
library(SpatialExperiment)

sce <- readH5AD("cosmx_nsclc.h5ad")

# Merge per-patient tumor subtype labels into a single "tumor" class
ct <- as.character(colData(sce)$cell_type)
ct[grepl("^tumor", ct)] <- "tumor"

spe <- SpatialExperiment(
  assays        = list(counts = assay(sce, "counts")),
  colData       = DataFrame(
    cell_type = ct,
    patient   = colData(sce)$patient,
    sample    = colData(sce)$sample
  ),
  spatialCoords = as.matrix(reducedDim(sce, "spatial"))
)
```

## Choosing `k_neighbors` with `plot_k_diagnostics()`

For abundant, well-distributed query populations (~150 K tumor cells
across 5 patients) the mean per-cell distance to the nearest tumor cell
stabilises immediately, and `k = 1` is adequate. For sparse query
populations (a few hundred cells per sample), larger `k` (3–5) reduces
noise. `plot_k_diagnostics()` gives a quick visual answer:

```{r k-diagnostics, eval = FALSE}
plot_k_diagnostics(
  input           = spe,
  query_celltype  = "tumor",
  celltype_column = "cell_type",
  sample_column   = "patient",
  k_range         = 1:10,
  max_distance_um = 200
)
```

If no clear elbow appears, the query is too sparse for robust per-cell
distance estimation; consider a coarser annotation or more replicates.

## Running RIPPLE

```{r run-ripple, eval = FALSE}
results <- run_ripple(
  input                = spe,
  query_celltype       = "tumor",
  celltype_column      = "cell_type",
  sample_column        = "patient",
  output_dir           = "ripple_output",
  analysis_name        = "tumor_ripple",
  k_neighbors          = 1,
  max_distance_um      = 200,
  min_cells_per_sample = 30,
  n_permutations       = 0,
  verbose              = TRUE
)
```

Key parameter choices:

- **`k_neighbors = 1`**: distance to the single nearest tumor cell;
  appropriate for paracrine signalling range.
- **`max_distance_um = 200`**: cells beyond 200 µm are excluded from the
  GLM. Captures typical secreted-factor gradients without long-range
  dilution.
- **`sample_column = "patient"`**: each patient is one biological
  replicate (N = 5). Per-patient coefficients are combined via Fisher's
  combined p-value with sign-consistency gating.
- **`n_permutations = 0`**: permutation testing skipped here; can be
  added later via `run_permutation_tests()` (or the GPU script in
  `inst/python/`).

## Loading cached results

The pre-computed outputs are bundled. In your own analysis, replace
these paths with files under your `output_dir/<analysis_name>/`.

```{r load-results}
results_path <- system.file("extdata/cosmx_cached/all_genes_results.csv",
                            package = "ripple")
results <- fread(results_path)

cat(nrow(results), "gene x cell_type combinations tested\n")
cat(sum(results$fisher_fdr < 0.05, na.rm = TRUE), "significant at Fisher FDR < 0.05\n")
```

## Pre-flight: contamination check (run this early!)

> **Strongly recommended:** run the contamination check **before** any
> downstream interpretation, top-hit listing, or pathway enrichment.
> Cross-cell-type "contamination" genes — genes that come up significant
> in many cell types simultaneously, typically ribosomal genes, MHC
> housekeeping, and ambient RNA from highly expressed query markers —
> will otherwise dominate every "top hits" table and every fGSEA leading
> edge, masking the cell-type-specific biology that RIPPLE is designed
> to surface.

`classify_gene_specificity()` returns one row per significant gene with
the number of cell types it appears in and a four-level class
(`specific`, `moderate`, `ubiquitous`, `contamination`). We compute it
**once**, here, and reuse it everywhere downstream.

```{r preflight-contam}
spec <- classify_gene_specificity(
  results,
  fdr_threshold           = 0.05,
  contamination_threshold = 4   # significant in >= 4 cell types -> contamination
)

table(spec$specificity_class)

# Genes flagged as contamination — to be deprioritised in top-hit lists
contam_genes <- spec[specificity_class == "contamination", unique(gene)]
cat("Contamination-flagged genes:", length(contam_genes), "\n")
head(spec[specificity_class == "contamination"][order(-n_celltypes)], 10)
```

```{r preflight-clean}
# Filtered copy used downstream when picking labels / top-hit lists.
# Plots that show *all* genes as background (volcanos, fGSEA) still use
# `results`; only the curated label / top-hit selections come from
# `results_clean`. This way a flagged gene like MIF is still visible as
# a dot on the volcano, but it does NOT crowd out the cell-type-specific
# hits in the labelled set.
results_clean <- results[!gene %in% contam_genes]
cat(nrow(results),       "rows in `results` (unfiltered)\n")
cat(nrow(results_clean), "rows in `results_clean` (contamination removed)\n")
```

> **⚠ Limitations of contamination filtering — read before reporting
> filtered results:**
>
> - **Loses real biology when contamination + induction co-occur.**
>   A gene can be both genuinely induced near the query *and* contaminated
>   by ambient RNA from query cells (MIF in this dataset is a textbook
>   example). Filtering removes the gene from every cell type's ranking,
>   including those where the induction is real. Always inspect the
>   contamination list and decide case-by-case.
> - **The threshold is panel-dependent.** Default
>   `contamination_threshold = 4` assumes ~10-20 cell types in the panel.
>   With 3 cell types the cutoff is unreachable; with 30 fine subtypes
>   it is too lenient. Re-tune for your annotation granularity.
> - **Removes genes globally, not per cell type.** A gene flagged
>   dataset-wide is dropped from every column. A gene that's
>   contamination across most cell types may still be a real signal in
>   the one cell type that biologically expresses it.
> - **Multi-cell-type biology is not contamination.** Cytokines, MHC
>   class II genes, and stress-response programs are legitimately
>   expressed across many cell types — the heuristic cannot tell them
>   apart from ambient RNA on its own.
> - **Rank-based tests are panel-size sensitive.** Dropping genes shifts
>   the ranks of all remaining genes; NES values from filtered and
>   unfiltered fGSEA runs are not directly comparable. Always report
>   whether the filter was applied and ideally show both.
>
> **Recommended workflow:** classify upfront (this section), keep both
> `results` (full) and `results_clean` (filtered) in scope, plot all
> genes as background but label / report top hits from `results_clean`,
> and run fGSEA both with and without `exclude_contamination = TRUE`
> (Panel D below).

# Reproducing paper Figure 4

The full multi-panel figure is rendered by `paper/render_figures.R`
(saves to `paper/figures/fig_4_cosmx.{pdf,png}`). Here we walk through
each panel with the same cached inputs.

## Panel A — Gradients per target cell type

```{r fig4-A, fig.width = 9, fig.height = 5}
plot_gene_counts_by_celltype(
  results,
  query_label             = "Tumor",
  contamination_threshold = 4
)
```

Each bar is split into two pieces by direction (induced near tumor vs
repressed near tumor) AND by specificity class — the faded portion is
the "ubiquitous / contamination" fraction (genes significant in ≥ 4
cell types). This separation is the key QC view: a tall bar is only
biologically interesting if its solid portion dominates. Fibroblasts
and macrophages are the most transcriptionally remodelled cell types
near tumor; both show a healthy specific-to-contamination ratio.

## Panel B — Fibroblast gradient volcano

The default `plot_gradient_volcano()` works well as a quick look:

```{r fig4-B-quick, fig.width = 8, fig.height = 6}
plot_gradient_volcano(
  results[cell_type == "fibroblast"],
  query_label = "Tumor",
  n_label     = 15
)
```

For the published figure we emphasise direction with red/blue colouring
and label a curated gene set. Note the labelling strategy: **all** genes
are still plotted as background dots (so the contamination-flagged
genes remain visible for interpretation), but the labelled set is
restricted to genes that are NOT in the contamination list, with a
couple of contamination genes (MIF, IGKC) kept explicitly so we can
discuss them.

```{r fig4-B-paper, fig.width = 8, fig.height = 6}
fib <- results[cell_type == "fibroblast"]    # all genes for the scatter
fib[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
fib[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Cell-type-specific top hits — pulled from `results_clean` so that
# ambient RNA does not crowd out the cell-type-specific biology.
top_clean_fib <- results_clean[
  cell_type == "fibroblast" & fisher_fdr < 0.05
][order(fisher_fdr)][1:12]$gene

# Plus a couple of contamination-flagged genes we want to discuss
# explicitly (see "Key fibroblast hits" below).
discuss_contam <- c("MIF", "IGKC", "JCHAIN", "RPL22")

label_fib <- unique(c(top_clean_fib, discuss_contam))
fib[, to_label := gene %in% label_fib & fisher_fdr < 0.05]

ggplot(fib, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = fib[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = 20, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = "Gradient score (per um)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Fibroblasts: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

**Key fibroblast hits (from `results_clean`):**

- **IGFBP5** (induced): IGF-binding protein involved in stromal
  remodelling; consistent with cancer-associated fibroblast (CAF)
  activation.
- **CXCL12** (repressed): chemokine normally expressed by quiescent
  fibroblasts; its repression near tumor reflects phenotypic switching
  to a CAF state.

**Contamination-flagged hits (kept on the plot, excluded from the
clean top-hit list):**

- **MIF** (induced): macrophage migration inhibitory factor — a
  well-known tumor-derived cytokine. *However*, MIF also appears as
  significant in 9 cell types and is flagged by the contamination
  check. This is the canonical "contamination + induction" case from
  the limitations note above: MIF is plausibly both a real fibroblast
  signal AND inflated by ambient RNA from highly MIF-expressing tumor
  cells. We still show it on the plot, but we did not let it compete
  for label slots with cell-type-specific hits.
- **IGKC / JCHAIN / RPL22**: immunoglobulin and ribosomal genes
  detected in fibroblasts at high significance — almost certainly
  ambient RNA from nearby plasma cells / leakage.

## Panel C — Per-patient consistency for MIF

A key feature of RIPPLE is per-replicate fitting. The forest plot below
shows per-patient gradient scores ± 95 % CI for MIF in fibroblasts:

```{r fig4-C, fig.width = 7, fig.height = 3.5}
coef_path <- system.file(
  "extdata/cosmx_cached/per_celltype/fibroblast/coef_per_sample.csv",
  package = "ripple"
)
coef_data <- fread(coef_path)

mif <- coef_data[gene == "MIF"]
mif[, ci_lo := coef - 1.96 * se]
mif[, ci_hi := coef + 1.96 * se]

ggplot(mif, aes(x = coef, y = sample_id, colour = sample_id)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi),
                orientation = "y", width = 0.25, alpha = 0.7) +
  geom_point(size = 3) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(
    x = "Per-patient gradient score for MIF (per um)",
    y = NULL,
    title = "MIF in fibroblasts: per-patient gradient scores"
  ) +
  theme_ripple(base_size = 12)
```

All five patients show a consistent negative coefficient — MIF RNA
counts increase near tumor in every replicate. Reproducibility alone
does not rule out systematic ambient RNA (both real induction and
ambient RNA can be reproducible across replicates), which is why the
contamination check below is a separate QC step.

## Panel D — Pathway enrichment (fGSEA dotplot)

```{r fig4-D-run, eval = FALSE}
# Run fGSEA via the atlas wrapper (writes fgsea_all_celltypes.csv)
run_ripple_atlas(
  results_dir = "ripple_output/tumor_ripple",
  query_label = "Tumor",
  run_fgsea   = TRUE,
  organism    = "human",
  fgsea_seed  = 42
)
```

```{r fig4-D-load}
fgsea_path <- system.file("extdata/cosmx_cached/fgsea_all_celltypes.csv",
                          package = "ripple")
fgsea_dt <- fread(fgsea_path)
```

`plot_fgsea_dotplot()` summarises the table at a glance — dot size
encodes -log10(padj), dot colour encodes NES on the same red/blue
diverging scale as the gradient volcano (red NES = induced near tumor,
blue NES = repressed), and non-significant cells are dimmed:

```{r fig4-D, fig.width = 9, fig.height = 7}
plot_fgsea_dotplot(fgsea_dt, padj_threshold = 0.1, top_n = 12)
```

Hallmark pathways like **E2F_TARGETS**, **G2M_CHECKPOINT** (proliferation,
red — induced in tumor-proximal fibroblasts and macrophages) and
**ALLOGRAFT_REJECTION** (immune effector programs, blue — depleted in
tumor-proximal myeloid + lymphoid cells) light up across multiple
stromal/immune cell types — each column tells its own story.

### Robust pathway enrichment: dropping contamination genes

The same contamination genes we filtered out of the volcano label sets
also dominate fGSEA leading edges (ribosomal genes, MHC housekeeping,
ambient-RNA-prone markers light up `INTERFERON_GAMMA_RESPONSE`,
`TNFA_SIGNALING_VIA_NFKB`, etc. in cell types where they don't really
belong). `run_ripple_fgsea()` accepts `exclude_contamination = TRUE` to
drop them before ranking — **using the same heuristic** as the
upfront classification above.

```{r fgsea-filtered, eval = FALSE}
# Using the same threshold we picked in the pre-flight section.
fgsea_filtered <- run_ripple_fgsea(
  results,
  gene_sets               = "hallmark",
  organism                = "human",
  exclude_contamination   = TRUE,
  contamination_threshold = 4
)

# Side-by-side: with and without filter
plot_fgsea_dotplot(fgsea_dt,       padj_threshold = 0.1, top_n = 12) +
  ggplot2::labs(title = "fGSEA -- all genes")
plot_fgsea_dotplot(fgsea_filtered, padj_threshold = 0.1, top_n = 12) +
  ggplot2::labs(title = "fGSEA -- contamination genes excluded")
```

When the panels are very similar, the unfiltered enrichment is robust
to the most obvious ambient RNA. When they differ substantially, the
filter is removing real signal, contamination, or both — investigate
the dropped leading-edge genes before reporting either version as the
answer. The full caveat list lives in the
[Pre-flight: contamination check] section above (and in
`?run_ripple_fgsea`); the most important one for fGSEA specifically is
that NES values are panel-size-sensitive and **not directly comparable
between filtered and unfiltered runs**.

## Panel E — Top tumor → fibroblast ligand–receptor pairs

Stage 6 (`run_ripple_lr()`) intersects gradient genes with curated
ligand–receptor databases and (optionally) NicheNet, producing a
combined prioritization score per L-R pair. The bundled file
`combined_prioritization.csv` is the output for fibroblasts:

```{r fig4-E-load}
lr_path <- system.file(
  "extdata/cosmx_cached/lr/fibroblast/combined_prioritization.csv",
  package = "ripple"
)
lr_fib <- fread(lr_path)
cat(nrow(lr_fib), "L-R pairs prioritized for tumor -> fibroblast\n")
head(lr_fib[order(-combined_score),
            .(ligand, receptor, combined_score, receptor_gradient_coef,
              receptor_fdr, ligand_pct_query)], 5)
```

```{r fig4-E, fig.width = 9, fig.height = 6}
lr_top <- lr_fib[order(-combined_score)][1:15]
lr_top[, pair := factor(paste0(ligand, " -> ", receptor),
                        levels = rev(paste0(ligand, " -> ", receptor)))]
lr_top[, direction := fifelse(receptor_gradient_coef < 0,
                              "Induced near tumor",
                              "Repressed near tumor")]

ggplot(lr_top, aes(x = combined_score, y = pair)) +
  geom_point(aes(size = ligand_pct_query, colour = direction)) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC"
  ), name = "Receptor gradient") +
  scale_size_continuous(range = c(2, 7),
                        name = "Tumor cells\nexpressing ligand (%)") +
  labs(
    x = "Combined prioritization score",
    y = NULL,
    title = "Top tumor -> fibroblast L-R pairs"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "right",
        legend.box = "vertical",
        legend.key.size = unit(0.4, "cm"))
```

The top pair (CD44 → EPCAM) and several MIF-axis pairs (MIF → CXCR4,
MIF → CD74) match the gradient hits in Panel B and the EMT pathway
enrichment in Panel D — three independent angles converging on the same
biology.

## Reproducing the full multi-panel figure

The exact PDF used in the paper is rendered by:

```{r render-fig4, eval = FALSE}
source("paper/render_figures.R")
# writes paper/figures/fig_4_cosmx.{pdf,png}
```

# Beyond Figure 4: macrophage deep-dive, contamination, and QC

## Macrophage gradient volcano

Macrophages show a polarised response near tumor (94 significant genes
in total), with both induced (antigen-presentation, cytoskeletal) and
repressed (homeostatic) programs. The same volcano view as Panel B,
applied to macrophages:

```{r mac-volcano, fig.width = 8, fig.height = 6}
mac <- results[cell_type == "macrophage"]   # all genes for the scatter
mac[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
mac[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Same labelling pattern as the fibroblast panel:
# top hits from `results_clean` (cell-type-specific signal) plus a
# curated set of biologically interesting macrophage markers (some of
# which — e.g. MHC II — are multi-cell-type by nature and therefore
# appear in `contam_genes`; we keep them because the macrophage
# induction is the biological point of interest).
top_clean_mac <- results_clean[
  cell_type == "macrophage" & fisher_fdr < 0.05
][order(fisher_fdr)][1:10]$gene

curated_mac <- c("HLA-DRA", "HLA-DPA1", "HLA-DRB1", "CD74",
                 "C1QA", "C1QB", "C1QC", "APOE",
                 "CXCL8", "IL1B", "CCL3", "MIF")

label_mac <- unique(c(top_clean_mac, curated_mac))
mac[, to_label := gene %in% label_mac & fisher_fdr < 0.05]

ggplot(mac, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = mac[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = 25, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = "Gradient score (per um)",
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Macrophages: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

**Macrophage induced near tumor (negative gradient score):**

- **HLA class II** (HLA-DRA, HLA-DPA1, HLA-DRB1) and **CD74** —
  antigen-presentation machinery. Tumor-proximal macrophages upregulate
  the MHC II axis, consistent with intratumoral antigen sampling.
- **Cytoskeletal remodelling** (TUBB, VIM, S100A4/A6) — supports
  motility and morphological adaptation in the tumor niche.
- **CXCL8 / IL1B / CCL3** — pro-inflammatory chemokines/cytokines
  characteristic of tumor-associated macrophage (TAM) activation.

**Macrophage repressed near tumor (positive gradient score):**

- **C1QA / C1QB / C1QC** and **APOE** — complement and lipid-handling
  programs of homeostatic / tissue-resident macrophages, downregulated
  as cells acquire a TAM phenotype.

This induced/repressed split — antigen-presentation/inflammatory **up**,
homeostatic/complement **down** — recovers the canonical macrophage
polarisation axis directly from spatial transcriptomics, anchored to
distance from the tumor.

# Specificity breakdown plot and QC dashboard

The contamination check itself was already run upfront (see
[Pre-flight: contamination check]); the `spec` object and `contam_genes`
vector are still in scope. `plot_specificity_breakdown()` summarises
the four classes as a one-glance bar:

```{r breakdown, fig.width = 7, fig.height = 4}
plot_specificity_breakdown(results, contamination_threshold = 4)
```

Genes like **HLA-E**, **KRT19**, **MIF**, and **RPL22** are flagged
because they are significant across 8–9 cell types — a hallmark of
ambient RNA. The MIF case (significant in 9 cell types, kept on the
fibroblast and macrophage volcanos for discussion) is the canonical
example of why filtering must be deliberate: MIF is plausibly both
genuinely induced by tumor paracrine signalling **and** inflated by
ambient RNA from high-expressing tumor cells.

## QC dashboard

The CosMx run was launched without a per-cell distance dump (older
version), so `ripple_plot_qc()` here will skip panels 1+2 and warn —
that's expected. On any fresh `run_ripple()` call all seven panels are
populated automatically.

```{r qc, eval = FALSE}
qc <- ripple_plot_qc(
  results_dir           = "ripple_output/tumor_ripple",
  query_signature_genes = c("KRT19", "EPCAM", "MKI67", "KRT8"),
  query_label           = "Tumor"
)
```

The bleed-through panel (panel 7) takes a query-marker vector; any
marker that shows up significant in many target cell types is
highlighted in red — the smoking gun for ambient RNA.

# Interpretation summary

This CosMx NSCLC analysis demonstrates RIPPLE's ability to detect
biologically interpretable spatial gradients in a multi-patient dataset:

1. **Fibroblasts** show the strongest tumor-proximity remodelling
   (129 genes), including CAF activation markers (IGFBP5, THBS1) and
   ECM genes. Top hits flagged as contamination (MIF, KRT19) should be
   validated independently.
2. **Macrophages** show a polarised response (94 genes), with
   tumor-proximity induction of antigen presentation (HLA-DRA,
   HLA-DPA1) and cytoskeletal remodelling (TUBB, S100A6).
3. **T cell subsets** (CD4 memory: 55 genes, CD8 memory: 27 genes,
   Tregs: 23 genes) each show distinct gradient signatures, suggesting
   spatial organisation of the adaptive immune response around tumor
   nests.
4. **Hallmark pathway enrichment** (EMT, hypoxia, TNFα–NFκB) recovers
   well-established tumor-niche biology and ties together
   stromal/immune programs.
5. **L-R prioritization** identifies CD44 → EPCAM and the MIF axis
   (MIF → CXCR4, MIF → CD74) as the top tumor → fibroblast signalling
   candidates — independently corroborated by the gradient and pathway
   panels.
6. **Contamination flagging** correctly identifies ambient RNA artefacts
   (immunoglobulin genes in non-immune cells, ubiquitous ribosomal
   genes) — critical for avoiding false biological conclusions.

All 516 significant genes are reproduced across 5 independent patients
with unanimous sign consistency (the strict gate), providing strong
evidence for genuine tumor-microenvironment spatial interactions.

## Session info

```{r session-info}
sessionInfo()
```
