---
title: "Getting started with RIPPLE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RIPPLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```

This vignette runs RIPPLE end-to-end on `ripple_mock_data`, the bundled
synthetic dataset. It takes under a minute. By the end you will have
run the core analysis, inspected the QC dashboard, and recovered the
planted gradient signal. The same workflow runs on real data. We use a
small toy example here so the vignette is fast to follow.

For a real-data walkthrough (CosMx NSCLC, ~765 K cells, 5 patients), see
`vignette("cosmx_nsclc_walkthrough", package = "ripple")`.

## What RIPPLE answers

> *For each gene in a target cell type, does expression change with
> physical distance to a chosen query cell type, reproducibly across
> biological replicates?*

RIPPLE is **not** a classic spatially-variable-gene test. It is **not** a
co-localisation test. It is **not** a ligand-receptor pairing. It has the defined goal of asking: does a gene's expression in a target cell type change with
distance to a query cell type, in the same direction in every replicate.

## Install and load

The bundled `ripple_mock_data` is a `SpatialExperiment`, so its Bioconductor
dependencies must be present before loading the package or this vignette.

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  "SpatialExperiment", "SingleCellExperiment", "SummarizedExperiment", "S4Vectors"
))
install.packages("R.utils") # for ripple_plot_qc() to read the .gz QC table

# Install from GitHub
remotes::install_github("Maier-Lab/RIPPLE")
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
```

## The bundled mock dataset

`ripple_mock_data` is a small `SpatialExperiment` with 50 genes × 600
cells across three samples. Three cell types: Tumor, T_cell, Fibroblast.
The gradient is planted in T cells only. 5 genes named `INDUCED_*` go
up near Tumor. 5 named `REPRESSED_*` go down. The other 40 (`BG_*`) are
flat.

```{r mock-data}
data(ripple_mock_data)
ripple_mock_data

table(
  sample   = ripple_mock_data$sample_id,
  celltype = ripple_mock_data$cell_type
)
```

### Using your own data

`run_ripple()` accepts a `Seurat`, `SingleCellExperiment`, or
`SpatialExperiment` object directly, or a path to an `.rds` file
containing one. If you are starting from raw matrices or CSV files,
build a canonical object first:

- `make_ripple_input(counts, metadata, coords, ...)` builds a
  `SpatialExperiment` (or `Seurat`) from in-memory matrices.
- `read_ripple_csv(dir_path, ...)` reads `counts.csv`, `metadata.csv`,
  and `coords.csv` from a directory and returns the same.

Spatial coordinates must be global (stitched across fields of view).
Per-FOV coordinates that reset to zero in each field will produce
meaningless distances. See `?make_ripple_input` for details.

## Run RIPPLE

`run_ripple()` is the main entry point. It auto-detects target cell
types (everything except the query), fits a per-sample Poisson GLM with
cell-size offset, and combines across replicates via Fisher's combined
p-value with sign-consistency gating.

```{r run-ripple, message = FALSE, warning = FALSE}
out_dir <- file.path(tempdir(), "ripple_quickstart")
unlink(out_dir, recursive = TRUE)
dir.create(out_dir)

results <- run_ripple(
  input                = ripple_mock_data,
  query_celltype       = "Tumor",
  celltype_column      = "cell_type",
  sample_column        = "sample_id",
  output_dir           = out_dir,
  min_cells_per_sample = 30,
  min_expr_pct         = 0,
  min_expr_floor       = 10,
  verbose              = FALSE
)

results_dir <- file.path(out_dir, "ripple")
list.files(results_dir, recursive = TRUE)
```

`run_ripple()` returns a `data.table` of results. It also writes an
output directory with per-cell-type CSVs, QC tables, and the per-cell
distance file that `ripple_plot_qc()` reads.

### One query vs one target cell type

To ask about a single target cell type (for example, only how `T_cell`
expression changes with distance from `Tumor`), pass a single value to
`target_celltypes`. This is faster and useful on large datasets or when
you only care about one interaction.

```{r one-vs-one, message = FALSE}
one_dir <- file.path(tempdir(), "ripple_one_vs_one")
unlink(one_dir, recursive = TRUE)
dir.create(one_dir)

# NOTE: running on a subset of targets makes RIPPLE warn that the
# cross-cell-type contamination check is incomplete -- see the caveat below.
tcell_only <- suppressWarnings(run_ripple(
  input                = ripple_mock_data,
  query_celltype       = "Tumor",
  celltype_column      = "cell_type",
  sample_column        = "sample_id",
  target_celltypes     = "T_cell",   # one query (Tumor) vs one target (T_cell)
  output_dir           = one_dir,
  min_cells_per_sample = 30,
  min_expr_pct         = 0,
  min_expr_floor       = 10,
  verbose              = FALSE
))

head(tcell_only[order(fisher_fdr), .(gene, cell_type, median_coef, fisher_fdr)])
```

**Caveat.** The cross-cell-type contamination heuristic (a gene
significant across many cell types is likely ambient RNA rather than a
real gradient) only works when RIPPLE has seen *all* cell types. On a
one-vs-one run it cannot flag broadly-contaminating genes, so
`run_ripple()` warns you. If contamination is a concern, run on all
targets, or supply known query markers via `query_signature_genes` (see
["Contamination is something you define"](#contamination-is-something-you-define-not-something-ripple-decides)
below).

## Look at the recovered signal

The planted `INDUCED_*` genes should appear with **negative** gradient
scores (RNA increases near Tumor) and pass FDR < 0.01. The `REPRESSED_*`
genes should be positive. Background genes (`BG_*`) should be roughly
uncalibrated noise.

```{r inspect-results}
head(results[, .(gene, cell_type, median_coef, fisher_fdr, sign_consistency)],
     5)

# Planted induced genes in T cells
results[
  grepl("^INDUCED", gene) & cell_type == "T_cell",
  .(gene, median_coef, fisher_fdr, sign_consistency)
]

# Planted repressed genes in T cells
results[
  grepl("^REPRESSED", gene) & cell_type == "T_cell",
  .(gene, median_coef, fisher_fdr, sign_consistency)
]
```

All ten planted genes should have the expected sign and full
sign-consistency (1.0) across the three samples.

## Volcano plot

```{r volcano, fig.height = 5}
plot_gradient_volcano(
  results[cell_type == "T_cell"],
  query_label = "Tumor",
  n_label     = 12
)
```

The 10 planted genes split cleanly: `INDUCED_*` on the left (negative
gradient score), `REPRESSED_*` on the right, background genes pile up
near the centre well below the FDR line.

## Specificity classification (candidate contamination)

`classify_gene_specificity()` separates cell-type-restricted signal from
genes that hit many cell types at once. Genes significant in at least
`broad_threshold` cell types are labelled `specificity_class == "broad"`.

```{r specificity}
spec <- classify_gene_specificity(
  results, fdr_threshold = 0.05, broad_threshold = 4
)
spec
```

In this synthetic dataset every planted gene comes back as `specific`,
significant in T_cell only. Real data is messier. You will see a mix of
`specific` (1 cell type), `moderate` (2 up to `broad_threshold` - 1 cell
types), and `broad` (at or above `broad_threshold`).

### Contamination is something you define, not something RIPPLE decides

You decide what counts as contamination. RIPPLE only proposes
**candidates**. It does not label any gene as "real" or "artefact" for
you. There are three candidate sources. None of them is ground truth:

- **Cross-cell-type heuristic.** `classify_gene_specificity()` flags
  broadly-significant genes (`specificity_class == "broad"`). A gene can
  be broad because it is genuine shared biology (e.g. a stress or
  interferon program active in several cell types), not because it is
  ambient RNA.
- **Ambient-family helpers.** `find_ambient_family_genes()` and
  `default_ambient_family_pattern()` return genes from families that
  commonly leak across segments (immunoglobulin, J-chain, ribosomal,
  mitochondrial). Membership in a family is a hint, not proof.
- **Query-marker spillover.** `query_signature_genes` is a list **you**
  supply of markers of the query cell type. Focal spillover of these
  markers into neighbours can be a strong contamination signal, yet such
  a gene may not be flagged as `broad` at all.

Treat each list as a starting point. Curate the actual contamination set
yourself using domain knowledge, and report what you removed and why.

Define your curated query-signature list once. Pass it to every
downstream step. The same genes get cleaned out at every stage. Use it
before pathway enrichment, ligand-receptor nomination, or any follow-up:

```r
# Your curated query-marker list (define from domain knowledge).
query_markers <- c("INDUCED_1", "INDUCED_2", "INDUCED_3")

# Pathway enrichment with the query signature removed before ranking.
run_ripple_fgsea(results, gene_sets = "hallmark",
                 query_signature_genes = query_markers)

# Ligand-receptor nomination drops receptors that are query markers
# (query-cell spillover, not real receiver receptors).
run_ripple_lr(results_dir, input = spe, query_celltype = "Tumor",
              celltype_column = "cell_type",
              query_signature_genes = query_markers)
```

Both `run_ripple_fgsea()` and `run_ripple_lr()` take
`query_signature_genes` for the same reason. Focal spillover often
shows up as `specific`, not `broad`. The cross-cell-type flag will not
catch it.

To make pathway enrichment ignore `broad`-class genes, pass
`exclude_broad = TRUE` to `run_ripple_fgsea()`. Read the "Limitations
of the broad-class filter" section in `?run_ripple_fgsea` first. Run
with and without the filter before reporting. The
[CosMx walkthrough](cosmx_nsclc_walkthrough.html) shows a side-by-side
comparison.

> **Note on the broad-class dials.** `classify_gene_specificity()`
> has two. `broad_threshold` sets how many cell types a gene must hit
> (default `4`, illustrative for a ~10-20 cell-type panel; tune for your
> data). `broad_sig_threshold` sets how strictly significant each "hit"
> must be (default `NULL`, which means any `*` hit counts; pass `"**"` or
> `"***"` to require stricter significance and flag fewer genes).
> **Pick values appropriate to your dataset and report them in the
> methods.** See `?classify_gene_specificity` for the full discussion.

## QC dashboard

`ripple_plot_qc()` builds a multi-panel figure from the files
`run_ripple()` writes. Two groups of panels.

Replicate quality:

- per-sample distance density
- cell composition
- sign consistency
- dispersion

Specificity:

- gene counts split by direction and specificity class
- specificity breakdown
- top widely-shared genes, optionally highlighted if they overlap your
  query-marker list

Query-marker bleed-through is a classic candidate pattern for
ambient-RNA contamination.

```{r qc-dashboard, fig.width = 14, fig.height = 18}
qc <- ripple_plot_qc(
  results_dir,
  query_signature_genes = c("INDUCED_1", "INDUCED_2", "INDUCED_3"),
  query_label           = "Tumor",
  top_n_bleed           = 10
)
qc
```

What to look for on real data:

- **Distance violins** (top-left): heavy right-skew or sample-specific
  spikes flag tissue-boundary effects.
- **Cell composition** (top-right): watch for sample dropouts or
  annotation imbalance. If one sample has 10× fewer T cells, its
  per-sample GLM will be noisy.
- **Sign consistency** (mid-left): genes with replicate-consistent
  direction pile up at 1.0; a flat distribution means most "hits" are
  driven by single samples and should not be trusted.
- **Median dispersion** (mid-right): Poisson is appropriate when
  cell-type medians sit between ~0.3 and ~2 (dashed lines). Cell types
  above 2 may need a quasi-Poisson model.
- **Bleed-through** (bottom): if a gene you passed in
  `query_signature_genes` shows up red AND is significant in many cell
  types, it is a strong candidate for ambient RNA from the query.

## Per-gene diagnostics

Two more plots help you inspect a single hit:

- `plot_gradient_curve()` draws the fitted expression-vs-distance curve
  for one gene in one cell type. It needs binned data from
  `bin_decay_data()`, which needs the raw counts and per-cell distances.
  Easiest to run while the input object is still in memory.
- `create_forest_plot()` shows per-sample coefficients and the combined
  estimate for one gene. Use it to see whether the gradient comes from
  every replicate or only one.

See the CosMx walkthrough for worked examples of both.

## Optional downstream stages

The core workflow above stands on its own: `run_ripple()`, the volcano,
the specificity check, the QC dashboard. The stages below are
**optional**. Run the ones that match your question:

- **Confounder control**, `run_ripple_confounder()`. Use when a second
  cell type shares the query's niche. Refits with distance to that
  control cell type added. Tells you whether a gradient is specific to
  the query or just reflects the shared neighbourhood.
- **Permutation testing**, `run_permutation_tests()` (or set
  `n_permutations` in `run_ripple()`). Builds an empirical null for
  query specificity. A GPU script is available for large datasets.
- **Pathway enrichment**, `run_ripple_fgsea()`. Summarises gradient
  genes as enriched pathways instead of a gene list.
- **Ligand-receptor integration**, `run_ripple_lr()`. Maps gradient
  genes to known ligand-receptor pairs that could mediate the effect.

## Next steps

- **Real data, deep dive:** `vignette("cosmx_nsclc_walkthrough", package = "ripple")`.
  765 K-cell, 5-patient CosMx NSCLC analysis. Covers the broad-class
  specificity check, fGSEA, ligand-receptor integration, and the full
  reproduction of paper Figure 3.
- **Statistical validation:** `vignette("benchmarks", package = "ripple")`.
  FDR calibration and power benchmarks.
- **Methods comparison:** `vignette("methods_positioning", package = "ripple")`.
  RIPPLE vs MISTy, nnSVG, CellChat, NicheNet.

## Session info

```{r session-info}
sessionInfo()
```
