---
title: "Getting started with RIPPLE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RIPPLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```

This vignette runs RIPPLE end-to-end on the bundled synthetic dataset
`ripple_mock_data` in under a minute. By the end you will have run the
core analysis, inspected the QC dashboard, and recovered the planted
gradient signal — the same workflow used on real data, just on a small
toy example so it's fast to follow.

For the deeper applied tutorial (real CosMx NSCLC, ~765 K cells, 5
patients), see `vignette("cosmx_nsclc_walkthrough", package = "ripple")`.

## What RIPPLE answers

> *For each gene in a target cell type, does expression change as a
> function of physical distance to a chosen query cell type —
> reproducibly across biological replicates?*

This is **not** a spatially-variable-gene test, **not** a co-localisation
test, and **not** a ligand–receptor pairing. It is a directed, cell-type
anchored, per-gene gradient test with built-in replicate-aware inference.
For comparisons against MISTy / nnSVG / CellChat / NicheNet, see
`vignette("methods_positioning", package = "ripple")`.

## Install and load

```{r install, eval = FALSE}
# From a private mirror or local source
remotes::install_github("CMangana/RIPPLE")
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
```

## The bundled mock dataset

`ripple_mock_data` is a small `SpatialExperiment` with 50 genes × 600
cells across three samples. It contains three cell types (Tumor,
T_cell, Fibroblast) and a planted distance-dependent gradient: 5 genes
named `INDUCED_*` and 5 named `REPRESSED_*` in T cells respond to
distance from Tumor; the remaining 40 background genes (`BG_*`) do not.

```{r mock-data}
data(ripple_mock_data)
ripple_mock_data

table(
  sample   = ripple_mock_data$sample_id,
  celltype = ripple_mock_data$cell_type
)
```

## Run RIPPLE

`run_ripple()` is the main entry point. It auto-detects target cell
types (everything except the query), fits a per-sample Poisson GLM with
cell-size offset, and combines across replicates via Fisher's combined
p-value with sign-consistency gating.

```{r run-ripple, message = FALSE, warning = FALSE}
out_dir <- file.path(tempdir(), "ripple_quickstart")
unlink(out_dir, recursive = TRUE)
dir.create(out_dir)

results <- run_ripple(
  input                = ripple_mock_data,
  query_celltype       = "Tumor",
  celltype_column      = "cell_type",
  sample_column        = "sample_id",
  output_dir           = out_dir,
  min_cells_per_sample = 30,
  min_expr_pct         = 0,
  min_expr_floor       = 10,
  verbose              = FALSE
)

results_dir <- file.path(out_dir, "ripple")
list.files(results_dir, recursive = TRUE)
```

`run_ripple()` returns a `data.table` of results AND writes a structured
output directory containing per-cell-type CSVs, QC tables, and the
per-cell distance file used by `ripple_plot_qc()`.

## Look at the recovered signal

The planted `INDUCED_*` genes should appear with **negative** gradient
scores (RNA increases near Tumor) and pass FDR < 0.01. The `REPRESSED_*`
genes should be positive. Background genes (`BG_*`) should be roughly
uncalibrated noise.

```{r inspect-results}
head(results[, .(gene, cell_type, median_coef, fisher_fdr, sign_consistency)],
     5)

# Planted induced genes in T cells
results[
  grepl("^INDUCED", gene) & cell_type == "T_cell",
  .(gene, median_coef, fisher_fdr, sign_consistency)
]

# Planted repressed genes in T cells
results[
  grepl("^REPRESSED", gene) & cell_type == "T_cell",
  .(gene, median_coef, fisher_fdr, sign_consistency)
]
```

All ten planted genes should have the expected sign and full
sign-consistency (1.0) across the three samples.

## Volcano plot

```{r volcano, fig.height = 5}
plot_gradient_volcano(
  results[cell_type == "T_cell"],
  query_label = "Tumor",
  n_label     = 12
)
```

The 10 planted genes split cleanly: `INDUCED_*` on the left (negative
gradient score), `REPRESSED_*` on the right, background genes pile up
near the centre well below the FDR line.

## Specificity classification

`classify_gene_specificity()` separates real cell-type-restricted signal
from genes that hit many cell types simultaneously — the latter are
candidate ambient RNA / segmentation contamination.

```{r specificity}
spec <- classify_gene_specificity(
  results, fdr_threshold = 0.05, contamination_threshold = 4
)
spec
```

In this clean synthetic dataset every planted gene shows up as
`specific` (significant in only one cell type — T_cell). On real data
you will typically see a mix of `specific`, `moderate`, and
`contamination` classes; the third is what you want to flag.

If you want pathway enrichment (`run_ripple_fgsea()`) to ignore
contamination-class genes, pass `exclude_contamination = TRUE` —
**but** read the "Limitations of contamination filtering" section in
`?run_ripple_fgsea` first, and ideally run with and without the filter
before reporting. The
[CosMx walkthrough](cosmx_nsclc_walkthrough.html) shows a side-by-side
comparison.

## QC dashboard

`ripple_plot_qc()` builds a multi-panel patchwork from the artefacts
written by `run_ripple()`. It bundles replicate-quality panels (per-sample
distance density, cell composition, sign consistency, dispersion) with
specificity / contamination panels (gene counts split by direction and
specificity, specificity breakdown, top widely-shared genes with optional
query-marker bleed-through highlighting).

```{r qc-dashboard, fig.width = 14, fig.height = 18}
qc <- ripple_plot_qc(
  results_dir,
  query_signature_genes = c("INDUCED_1", "INDUCED_2", "INDUCED_3"),
  query_label           = "Tumor",
  top_n_bleed           = 10
)
qc
```

What to look for on real data:

- **Distance violins** (top-left): heavy right-skew or sample-specific
  spikes flag tissue-boundary effects.
- **Cell composition** (top-right): watch for sample dropouts or
  annotation imbalance — if one sample has 10× fewer T cells, its
  per-sample GLM will be noisy.
- **Sign consistency** (mid-left): genes with replicate-consistent
  direction pile up at 1.0; a flat distribution means most "hits" are
  driven by single samples and should not be trusted.
- **Median dispersion** (mid-right): Poisson is appropriate when
  cell-type medians sit between ~0.3 and ~2 (dashed lines). Cell types
  above 2 may need a quasi-Poisson model.
- **Bleed-through** (bottom): if a gene you passed in
  `query_signature_genes` shows up red AND is significant in many cell
  types, it is almost certainly ambient RNA from the query.

## Next steps

- **Real data, deep dive:** `vignette("cosmx_nsclc_walkthrough", package = "ripple")`
  walks through a 765 K-cell, 5-patient CosMx NSCLC analysis — including
  contamination handling, fGSEA pathway enrichment, ligand–receptor
  integration, and full reproduction of paper Figure 4.
- **Statistical validation:** `vignette("benchmarks", package = "ripple")`
  shows the FDR calibration and power benchmarks that back the method.
- **Methods comparison:** `vignette("methods_positioning", package = "ripple")`
  compares RIPPLE against MISTy, nnSVG, CellChat, NicheNet.

## Session info

```{r session-info}
sessionInfo()
```
