---
title: "RIPPLE in the spatial transcriptomics methods landscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE in the spatial transcriptomics methods landscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

## The question RIPPLE answers

Many spatial transcriptomics tools detect spatially variable genes, infer
cell-cell communication, or identify tissue niches. RIPPLE asks a different
question:

> **For each gene in a target cell type, does expression change as a function
> of physical distance to a chosen query cell type, reproducibly across
> biological replicates?**

This is not a spatially variable gene test. It is not a
ligand-receptor pairing. It is a directional, per-gene test of
expression vs distance to one specific cell type, with the biological
replicate as the unit of inference.

## Methods landscape

The table below covers tools that a user might reach for when
answering a similar-sounding question about gene expression in a
spatial context. Cell-cell communication tools (CellChat, COMMOT,
NicheNet) are not included: they answer a different question (which
ligand-receptor pairs are active between cell types) and are
complementary to RIPPLE rather than alternatives. RIPPLE's own Stage 6
(`run_ripple_lr`) integrates L-R priors downstream to add mechanism to
gradient hits.

```{r landscape-table}
landscape <- data.frame(
  Tool = c(
    "nnSVG / SpatialDE / SPARK / Hotspot",
    "STdiff (spatialGE)",
    "MISTy",
    "BANKSY / CellCharter",
    "RIPPLE"
  ),
  Question = c(
    "Spatially variable gene / module detection",
    "Differential expression between predefined groups",
    "Per-gene neighbourhood importance (multi-view ML)",
    "Spatial niche / domain clustering",
    "Distance-conditioned gene gradients per replicate"
  ),
  Output = c(
    "Per-gene p-value (omnibus, non-directional)",
    "Per-group contrast (logFC + p-value)",
    "Variable importance (default: random forest)",
    "Cluster / niche labels",
    "Gradient score + Fisher FDR + decay slope"
  ),
  Unit = c(
    "Per section",
    "Per section",
    "Per section",
    "Per section or joint across sections",
    "Biological replicate (Fisher-combined)"
  ),
  Directional = c(
    "No",
    "Yes (group contrast)",
    "Indirect (via partial dependence)",
    "N/A",
    "Yes (sign of gradient score)"
  ),
  Celltype_anchored = c(
    "No",
    "Group labels (any partition)",
    "Configurable",
    "N/A",
    "Yes (query anchor)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(landscape, col.names = c(
  "Tool", "Question answered", "Output type",
  "Unit of inference", "Directional?", "Cell-type anchored?"
), caption = paste(
  "Comparison of spatial transcriptomics tools that address similar",
  "questions to RIPPLE. 'Unit of inference' is the level at which each",
  "tool is designed to compute its primary statistic. Tools listed as",
  "'Per section' are statistically valid within one section;",
  "multi-sample aggregation is left to the user."
))
```

## What makes RIPPLE distinct

**1. Replicate-aware by design.**
Most spatial transcriptomics tools were designed to analyse one tissue
section at a time. Statistically sound for that scope, but multi-sample
aggregation is left to the user, and naive pooling of cells across
samples inflates the effective sample size and can drive
pseudoreplication. RIPPLE treats each biological replicate (patient
or mouse) as the unit of inference from the start: it fits a per-sample
Poisson GLM, then combines coefficients across replicates via Fisher's
combined p-value plus a sign-consistency gate. Only gradients that hold
across replicates pass.

**2. Directional, interpretable output.**
Each gene gets a signed **gradient score**, the median of its
per-sample log-rate coefficients (log-rate change per micrometre). A
negative score means the gene is *induced* near the query cell type. A
positive score means it is *repressed*. The score reads directly as a
spatial gradient. Importance scores (MISTy) and omnibus p-values
(nnSVG, SPARK) do not.

**3. Cell-type anchored.**
Distance is always to a specific query cell type. For example,
"distance to the nearest tumour cell". SVG methods catch every
spatially structured pattern, including ones unrelated to the query
(tissue architecture, technical gradients). RIPPLE answers the
biological question only.

**4. Cell-size offset.**
The Poisson GLM uses `offset(log(total_counts))` per cell. This
converts raw counts to rates. It removes a known confound in imaging
platforms: cells near the query can look bigger or more
transcript-rich, and the offset corrects for that.

## Closest comparator: MISTy

[MISTy](https://doi.org/10.1186/s13059-022-02663-5) (Tanevski et al., 2022)
is the most conceptually adjacent tool. With its celltype-aware para-view,
MISTy can model gene expression as a function of neighbourhood composition,
including proximity to a specific cell type. Key differences:

| | MISTy | RIPPLE |
|---|---|---|
| **Statistical framework** | Multi-view ML (random forest by default, linear also supported) | Poisson GLM with cell-size offset |
| **Output** | Importance score (unitless) | Gradient score (log-rate change per µm) |
| **Direction** | Not directly available (recoverable via partial dependence) | Signed (induced vs repressed) |
| **Replicate handling** | Per-section views; pool or aggregate importances across samples | Per-sample GLM; Fisher's combined p + sign-consistency gate |
| **FDR control** | Not built in (importance scores) | Validated under null simulation |

MISTy and RIPPLE are complementary. MISTy is exploratory: it discovers
which neighbourhood features *matter*. RIPPLE is confirmatory: it tests
a specific query-target hypothesis with directional, replicate-aware
inference.

## Continuous distance vs. group contrast: STdiff

[STdiff](https://doi.org/10.1186/s13059-024-03379-4) (Ospina et al.,
2024, part of spatialGE) tests differential expression between
predefined groups of cells in a spatial context. Groups can be tissue
domains, cell clusters, or phenotype labels. STdiff offers two
testing modes that account for within-section spatial autocorrelation:
a non-parametric spatial permutation test and a parametric spatial
mixed model. Both compare expression between groups.

The form of the spatial variable is the key difference from RIPPLE.
STdiff asks "does gene X differ between group A and group B?", a
categorical contrast. RIPPLE asks "does gene X change continuously
with distance to the query cell type?", a continuous gradient. You
could approximate the RIPPLE question by binning cells into near vs
far and running a group test. That loses the gradient shape and the
decay slope that RIPPLE estimates directly.

The two are complementary rather than competing. Use STdiff when the
biological hypothesis is a group contrast (region A vs region B,
cluster X vs cluster Y). Use RIPPLE when the hypothesis is a continuous
distance-dependent gradient from a specific query cell type.

## When to use RIPPLE vs. alternatives

- **"Which genes are spatially variable?"** → nnSVG, SpatialDE, SPARK
- **"Which genes differ between two tissue domains or clusters?"** → STdiff
- **"Which L-R pairs drive communication?"** → CellChat, COMMOT, NicheNet
- **"What neighbourhood features predict gene X?"** → MISTy
- **"Which genes in cell type B change with distance to cell type A, reproducibly across patients?"** → **RIPPLE**

## Optional downstream stages

The core RIPPLE result is the per-gene gradient score with its Fisher-combined
FDR. Several stages are optional and are run only when the analysis calls for
them:

- **Confounder control** (`run_ripple_confounder`) adds distance to a control
  cell type, to separate query-specific gradients from shared-niche effects.
- **Permutation testing** (`run_permutation_tests`) builds a label-shuffled
  null to confirm query specificity.
- **fGSEA** (`run_ripple_fgsea`) summarises gradient genes into pathways.
- **L-R integration** (`run_ripple_lr`) matches gradient genes to known
  ligand-receptor pairs, adding mechanistic context to the statistical
  gradients.

Opt into each stage based on the question and the controls you have.
None of them is required to obtain gradient scores. The getting-started
and walkthrough vignettes show when and how to add them.

## A note on contamination

RIPPLE does not decide what counts as contamination. It proposes
candidates. Three sources:

- `classify_gene_specificity()` flags genes significant near many
  unrelated cell types (`specificity_class == "broad"`).
- `find_ambient_family_genes()` flags common ambient families
  (immunoglobulin, J-chain, ribosomal, mitochondrial).
- A `query_signature_genes` list (yours) flags query-marker
  bleed-through.

You curate the final list with domain knowledge. See the
getting-started and `cosmx_nsclc_walkthrough` vignettes for the
workflow.

## See also

- `vignette("getting_started", package = "ripple")`: 5-minute end-to-end
  on the bundled synthetic dataset.
- `vignette("cosmx_nsclc_walkthrough", package = "ripple")`: applied
  case study reproducing paper Figure 4 (CosMx NSCLC, 5 patients).
- `vignette("benchmarks", package = "ripple")`: FDR calibration, power
  curves, and runtime measurements (paper Figure S1).

## References

- Svensson et al. (2018). SpatialDE: identification of spatially variable
  genes. *Nature Methods*.
- Sun et al. (2020). Statistical analysis of spatial expression patterns
  for spatially resolved transcriptomic studies (SPARK). *Nature Methods*.
- Weber et al. (2023). nnSVG for the scalable identification of spatially
  variable genes using nearest-neighbor Gaussian processes. *Nature Methods*.
- DeTomaso & Yosef (2021). Hotspot identifies informative gene modules
  across modalities of single-cell genomics. *Cell Systems*.
- Ospina et al. (2024). spatialGE provides modeling and visualization tools
  for spatial transcriptomics data, including STdiff for spatially informed
  differential expression. *Genome Biology*.
- Tanevski et al. (2022). Explainable multiview framework for dissecting
  spatial relationships from highly multiplexed data. *Genome Biology*.
- Cang et al. (2023). Screening cell-cell communication in spatial
  transcriptomics via collective optimal transport (COMMOT). *Nature Methods*.
- Jin et al. (2021). Inference and analysis of cell-cell communication using
  CellChat. *Nature Communications*.
- Browaeys et al. (2020). NicheNet: modeling intercellular communication by
  linking ligands to target genes. *Nature Methods*.
- Singhal et al. (2024). BANKSY unifies cell typing and tissue domain
  segmentation for scalable spatial omics data analysis. *Nature Genetics*.
- Varrone et al. (2024). CellCharter reveals spatial cell niches associated
  with tissue remodeling and cell plasticity. *Nature Genetics*.
