---
title: "RIPPLE in the spatial transcriptomics methods landscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE in the spatial transcriptomics methods landscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

## The question RIPPLE answers

Many spatial transcriptomics tools detect spatially variable genes, infer
cell-cell communication, or identify tissue niches. RIPPLE asks a different
question:

> **For each gene in a target cell type, does expression change as a function
> of physical distance to a chosen query cell type, reproducibly across
> biological replicates?**

This is not an SVG test. It is not a ligand-receptor pairing. It is a
directional, per-gene test of expression vs distance to one specific
cell type, with replicate-aware inference built in.

## Methods landscape

```{r landscape-table}
landscape <- data.frame(
  Tool = c(
    "nnSVG / SpatialDE / SPARK", "STdiff (spatialGE)",
    "MISTy", "COMMOT / CellChat",
    "NicheNet", "BANKSY / CellCharter",
    "RIPPLE"
  ),
  Question = c(
    "Spatially variable gene detection",
    "Differential expression between discrete groups",
    "Per-gene neighbourhood importance (multi-view ML)",
    "Cell-cell communication via L-R optimal transport",
    "Ligand activity prediction from target gene expression",
    "Spatial niche / domain clustering",
    "Distance-conditioned gene gradients per replicate"
  ),
  Output = c(
    "p-value (omnibus, non-directional)",
    "Per-group contrast (logFC, p-value)",
    "Variable importance (RF)",
    "Communication score",
    "Ligand activity score",
    "Cluster labels",
    "Gradient score + Fisher FDR + decay slope"
  ),
  Replicate_aware = c("No", "No", "No", "No", "No", "No", "Yes"),
  Directional = c(
    "No", "Group contrast", "Indirect", "L to R", "L to R", "N/A",
    "Yes (sign of gradient score)"
  ),
  Celltype_anchored = c(
    "No", "Group labels", "Configurable", "Cell-cell pairs",
    "Sender-receiver", "N/A", "Yes (query anchor)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(landscape, col.names = c(
  "Tool", "Question answered", "Output type",
  "Replicate-aware?", "Directional?", "Cell-type anchored?"
), caption = "Comparison of spatial transcriptomics analysis approaches.")
```

## What makes RIPPLE distinct

**1. Replicate-aware inference.**
Most spatial tools treat each cell as a unit of replication. RIPPLE
treats each biological sample (patient or mouse) as the unit. It fits a
per-sample Poisson GLM and combines across replicates with Fisher's
combined p-value plus a sign-consistency gate. This avoids
pseudoreplication. Only gradients that hold across replicates pass.

**2. Directional, interpretable output.**
Each gene gets a signed **gradient score**, the median of its
per-sample log-rate coefficients (log-rate change per micrometre). A
negative score means the gene is *induced* near the query cell type. A
positive score means it is *repressed*. The score reads directly as a
spatial gradient. Importance scores (MISTy) and omnibus p-values
(nnSVG, SPARK) do not.

**3. Cell-type anchored.**
Distance is always to a specific query cell type. For example,
"distance to the nearest tumour cell". SVG methods catch every
spatially structured pattern, including ones unrelated to the query
(tissue architecture, technical gradients). RIPPLE answers the
biological question only.

**4. Cell-size offset.**
The Poisson GLM uses `offset(log(total_counts))` per cell. This
converts raw counts to rates. It removes a known confound in imaging
platforms: cells near the query can look bigger or more
transcript-rich, and the offset corrects for that.

## Closest comparator: MISTy

[MISTy](https://doi.org/10.1186/s13059-022-02663-5) (Tanevski et al., 2022)
is the most conceptually adjacent tool. With its celltype-aware para-view,
MISTy can model gene expression as a function of neighbourhood composition,
including proximity to a specific cell type. Key differences:

| | MISTy | RIPPLE |
|---|---|---|
| **Statistical framework** | Random Forest variable importance | Poisson GLM with offset |
| **Output** | Importance score (unitless) | Gradient score (log-rate change per µm) |
| **Direction** | Not directly available | Signed (induced vs. repressed) |
| **Replicate handling** | Pool or run per-sample independently | Fisher's combined p + sign-consistency gate |
| **FDR control** | Not formally calibrated | Validated under null simulation |

MISTy and RIPPLE are complementary. MISTy is exploratory: it discovers
which neighbourhood features *matter*. RIPPLE is confirmatory: it tests
a specific query-target hypothesis with directional, replicate-aware
inference.

## Continuous distance vs. group contrast: STdiff

[STdiff](https://doi.org/10.1186/s13059-024-03379-4) (Ospina et al.,
2024, part of spatialGE) tests differential expression between discrete
groups of cells. A group is a tissue domain or a cell cluster /
phenotype. STdiff fits spatial linear mixed models with an exponential
covariance to handle within-sample spatial autocorrelation. It then
contrasts expression between groups. It does not model expression as a
continuous function of distance.

The form of the spatial variable is the key difference. STdiff asks
"does gene X differ between group A and group B?". A categorical
contrast. RIPPLE asks "does gene X change continuously with distance
to the query cell type?". A continuous gradient. You could approximate
the RIPPLE question by binning cells near vs far. That loses the
gradient shape and the decay slope that RIPPLE estimates directly.

## When to use RIPPLE vs. alternatives

- **"Which genes are spatially variable?"** → nnSVG, SpatialDE, SPARK
- **"Which genes differ between two tissue domains or clusters?"** → STdiff
- **"Which L-R pairs drive communication?"** → CellChat, COMMOT, NicheNet
- **"What neighbourhood features predict gene X?"** → MISTy
- **"Which genes in cell type B change with distance to cell type A, reproducibly across patients?"** → **RIPPLE**

## Optional downstream stages

The core RIPPLE result is the per-gene gradient score with its Fisher-combined
FDR. Several stages are optional and are run only when the analysis calls for
them:

- **Confounder control** (`run_ripple_confounder`) adds distance to a control
  cell type, to separate query-specific gradients from shared-niche effects.
- **Permutation testing** (`run_permutation_tests`) builds a label-shuffled
  null to confirm query specificity.
- **fGSEA** (`run_ripple_fgsea`) summarises gradient genes into pathways.
- **L-R integration** (`run_ripple_lr`) matches gradient genes to known
  ligand-receptor pairs, adding mechanistic context to the statistical
  gradients.

Opt into each stage based on the question and the controls you have.
None of them is required to obtain gradient scores. The getting-started
and walkthrough vignettes show when and how to add them.

## A note on contamination

RIPPLE does not decide what counts as contamination. It proposes
candidates. Three sources:

- `classify_gene_specificity()` flags genes significant near many
  unrelated cell types (`specificity_class == "broad"`).
- `find_ambient_family_genes()` flags common ambient families
  (immunoglobulin, J-chain, ribosomal, mitochondrial).
- A `query_signature_genes` list (yours) flags query-marker
  bleed-through.

You curate the final list with domain knowledge. See the
getting-started and `cosmx_nsclc_walkthrough` vignettes for the
workflow.

## See also

- `vignette("getting_started", package = "ripple")`: 5-minute end-to-end
  on the bundled synthetic dataset.
- `vignette("cosmx_nsclc_walkthrough", package = "ripple")`: applied
  case study reproducing paper Figure 4 (CosMx NSCLC, 5 patients).
- `vignette("benchmarks", package = "ripple")`: FDR calibration, power
  curves, and runtime measurements (paper Figure 2).

## References

- Weber et al. (2023). nnSVG for the scalable identification of spatially
  variable genes using nearest-neighbor Gaussian processes. *Nature Methods*.
- Ospina et al. (2024). spatialGE provides modeling and visualization tools
  for spatial transcriptomics data, including STdiff for spatially informed
  differential expression. *Genome Biology*.
- Tanevski et al. (2022). Explainable multiview framework for dissecting
  spatial relationships from highly multiplexed data. *Genome Biology*.
- Cang et al. (2023). Screening cell-cell communication in spatial
  transcriptomics via collective optimal transport. *Nature Methods*.
- Browaeys et al. (2020). NicheNet: modeling intercellular communication by
  linking ligands to target genes. *Nature Methods*.
- Singhal et al. (2024). BANKSY unifies cell typing and tissue domain
  segmentation for scalable spatial omics data analysis. *Nature Genetics*.
