---
title: "RIPPLE in the spatial transcriptomics methods landscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE in the spatial transcriptomics methods landscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

## The question RIPPLE answers

Many spatial transcriptomics tools detect spatially variable genes, infer
cell--cell communication, or identify tissue niches. RIPPLE asks a distinct
question:

> **For each gene in a target cell type, does expression change as a function
> of physical distance to a chosen query cell type --- reproducibly across
> biological replicates?**

This is neither a general spatially-variable-gene (SVG) test nor a
ligand--receptor pairing. It is a **directed, cell-type-anchored, per-gene
gradient test** with built-in replicate-aware inference.

## Methods landscape

```{r landscape-table}
landscape <- data.frame(
  Tool = c(
    "Hotspot", "nnSVG / SpatialDE2 / SPARK",
    "MISTy", "COMMOT / CellChat",
    "NicheNet", "BANKSY / CellCharter",
    "RIPPLE"
  ),
  Question = c(
    "Spatially autocorrelated gene modules",
    "Spatially variable gene detection",
    "Per-gene neighbourhood importance (multi-view ML)",
    "Cell-cell communication via L-R optimal transport",
    "Ligand activity prediction from target gene expression",
    "Spatial niche / domain identification",
    "Distance-conditioned gene gradients per replicate"
  ),
  Output = c(
    "Z-score, gene modules",
    "p-value (omnibus)",
    "Variable importance (RF)",
    "Communication score",
    "Ligand activity score",
    "Cluster labels",
    "Coefficient + Fisher FDR + decay slope"
  ),
  Replicate_aware = c("No", "No", "No", "No", "No", "No", "Yes"),
  Directional = c(
    "No", "No", "Indirect", "L to R", "L to R", "N/A",
    "Yes (sign of beta)"
  ),
  Celltype_anchored = c(
    "No", "No", "Configurable", "Cell-cell pairs",
    "Sender-receiver", "N/A", "Yes (query anchor)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(landscape, col.names = c(
  "Tool", "Question answered", "Output type",
  "Replicate-aware?", "Directional?", "Cell-type anchored?"
), caption = "Comparison of spatial transcriptomics analysis approaches.")
```

## What makes RIPPLE distinct

**1. Replicate-aware inference.**
Most spatial tools operate on a single sample. RIPPLE fits a per-sample
Poisson GLM and combines across biological replicates using Fisher's combined
p-value with a sign-consistency gate. This avoids pseudoreplication (treating
cells as independent observations) and ensures that reported gradients are
reproducible across patients or mice --- not artefacts of a single tissue
section.

**2. Directional, interpretable output.**
Each gene receives a signed coefficient (beta, in log-rate change per
micrometre). A negative beta means gene expression *increases* near the query
cell type (induced); a positive beta means it *decreases* (repressed). This is
directly interpretable as a spatial gradient, unlike importance scores (MISTy)
or omnibus p-values (nnSVG).

**3. Cell-type-anchored.**
The distance metric is always to a specific query cell type (e.g., "distance
to the nearest tumour cell"). SVG methods detect all spatially structured
expression, including genes that are spatially variable for reasons unrelated
to the query (tissue architecture, technical gradients). RIPPLE is specific to
the biological question.

**4. Cell-size offset.**
The Poisson GLM includes `offset(log(total_counts))` per cell, converting raw
counts to rates. This controls for ambient RNA and segmentation artefacts that
co-vary with cell size --- a critical confound in imaging-based platforms where
cells near the query may appear larger or more transcript-rich.

## Closest comparator: MISTy

[MISTy](https://doi.org/10.1186/s13059-022-02663-5) (Tanevski et al., 2022)
is the most conceptually adjacent tool. With its celltype-aware para-view,
MISTy can model gene expression as a function of neighbourhood composition,
including proximity to a specific cell type. Key differences:

| | MISTy | RIPPLE |
|---|---|---|
| **Statistical framework** | Random Forest variable importance | Poisson GLM with offset |
| **Output** | Importance score (unitless) | Gradient score (log-rate change per µm) |
| **Direction** | Not directly available | Signed (induced vs. repressed) |
| **Replicate handling** | Pool or run per-sample independently | Fisher's combined p + sign-consistency gate |
| **FDR control** | Not formally calibrated | Validated under null simulation |

MISTy and RIPPLE are **complementary**: MISTy excels at discovering which
neighbourhood features *matter* (exploratory), while RIPPLE provides
**calibrated, directional, replicate-aware inference** for a specific
query-target hypothesis (confirmatory).

## When to use RIPPLE vs. alternatives

- **"Which genes are spatially variable?"** → nnSVG, SpatialDE2, SPARK, Hotspot
- **"Which L-R pairs drive communication?"** → CellChat, COMMOT, NicheNet
- **"What neighbourhood features predict gene X?"** → MISTy
- **"Which genes in cell type B change with distance to cell type A, reproducibly across patients?"** → **RIPPLE**

RIPPLE's optional Stage 6 (L-R integration) bridges back to ligand--receptor
tools by matching gradient genes to known L-R pairs, providing mechanistic
context for the statistical gradients.

## References

- DeTomaso & Yosef (2021). Hotspot identifies informative gene modules across
  modalities of single-cell genomics. *Cell Systems*.
- Weber et al. (2023). nnSVG for the scalable identification of spatially
  variable genes using nearest-neighbor Gaussian processes. *Nature Methods*.
- Tanevski et al. (2022). Explainable multiview framework for dissecting
  spatial relationships from highly multiplexed data. *Genome Biology*.
- Cang et al. (2023). Screening cell--cell communication in spatial
  transcriptomics via collective optimal transport. *Nature Methods*.
- Browaeys et al. (2020). NicheNet: modeling intercellular communication by
  linking ligands to target genes. *Nature Methods*.
- Singhal et al. (2024). BANKSY unifies cell typing and tissue domain
  segmentation for scalable spatial omics data analysis. *Nature Genetics*.
