---
title: "Parallelising RIPPLE on large datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallelising RIPPLE on large datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`run_ripple()` runs on one core. On the Xenium 5K panel with ~70,000
cells and ~7 cell types, that can mean several hours. If you have a
multi-core machine, you can fan out over cell types externally
without changing anything in RIPPLE itself.

In summary, subset the input to `query + one target` per cell type,
dispatch each subset to its own worker with `future.apply`, and merge
the results.

## When to use this

- The dataset is large (thousands of cells per cell type, hundreds to
  thousands of genes).
- You have more than one CPU core available.
- You do not have SLURM. If you do, use the array-job scripts in
  `inst/slurm/` instead. Cluster fan-out is simpler than in-process
  parallelism.

When **not** to use this:

- Small datasets. Parallel overhead beats the win.
- Memory-tight systems. Each worker duplicates the input object on
  Windows (multisession cannot fork).

## The pattern

Install `future.apply` if you do not have it:

```r
install.packages("future.apply")
```

Set up the parallel plan. Pick `workers` based on cores and memory:

```r
library(ripple)
library(future.apply)
library(data.table)

plan(multisession, workers = 4)
```

Load the object once:

```r
xen <- readRDS("xenium_combined.rds")
```

Fan out over target cell types. Each worker gets a subset with only
`query + one target`, runs a full `run_ripple()` for that pair, and
writes to its own output subdirectory so there are no collisions:

```r
query <- "endothelial"
target_types <- setdiff(unique(xen$annotation), query)
out_root <- "ripple_out_parallel"

results_list <- future_lapply(target_types, function(ct) {
  keep <- xen$annotation %in% c(query, ct)
  xen_sub <- xen[, keep]

  ripple::run_ripple(
    input                = xen_sub,
    query_celltype       = query,
    celltype_column      = "annotation",
    sample_column        = "sample",
    target_celltypes     = ct,
    output_dir           = file.path(out_root, ct),
    analysis_name        = "ripple",
    min_cells_per_sample = 30,
    min_expr_pct         = 0,
    min_expr_floor       = 10,
    verbose              = FALSE
  )
}, future.seed = TRUE)

names(results_list) <- target_types
```

`future.seed = TRUE` silences the "non-reproducible RNG" warning.
RIPPLE calls `set.seed(42)` internally, so per-worker seeding is
deterministic either way.

## Merge the per-cell-type results

For a quick combined data.table, `rbindlist` is enough:

```r
combined <- rbindlist(results_list, fill = TRUE)
```

`classify_gene_specificity`, `run_ripple_fgsea`, and
`plot_gradient_volcano` accept this combined data.table directly.

`ripple_plot_qc()` is the exception. It expects a full results
directory (per_celltype/, qc/, summary/) rather than a data.table, and
the parallel fan-out fragments that tree across worker directories.
Use `consolidate_parallel_ripple()` to merge the workers into a single
canonical results tree:

```r
worker_dirs <- file.path(out_root, target_types, "ripple")

final_dir <- consolidate_parallel_ripple(
  worker_dirs    = worker_dirs,
  output_dir     = file.path(out_root, "combined"),
  query_celltype = query
)
```

The consolidation copies each worker's per_celltype/ subtree, merges
`cell_distances.csv.gz` across workers (deduping the query cells so
they are not counted N times), and rebuilds the combined summary via
`merge_ripple_results()`. Now `ripple_plot_qc()` and any other
directory-based downstream function work as if the run had been
serial:

```r
ripple_plot_qc(
  results_dir           = final_dir,
  query_signature_genes = c("EPCAM", "KRT8"),  # your query markers
  query_label           = query
)
```

## Memory notes

The `multisession` backend copies the R session to each worker. With
`workers = 4` and a 500 MB Seurat object, expect ~2 GB extra RAM.
Reduce `workers` if the machine starts to swap.

On Linux and macOS you can pick `multicore` instead of `multisession`.
It uses `fork()` and shares memory copy-on-write, so the extra RAM is
much smaller. Windows does not have `fork()`; `multicore` falls back
to sequential there, so use `multisession` on Windows.

## Reset the plan when done

```r
plan(sequential)
```

## Alternative: SLURM array jobs

If you have a cluster, use `inst/slurm/run_hymy_distance_correlation_v2.sh`
and the associated array-driven scripts. One SLURM task per cell
type, one core per task, no `future.apply` involvement. Same fan-out
shape as this vignette, but at the scheduler level. This is what we
used for the CosMx and HyMy paper analyses.

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
